---
title: "Week 10 - Writing Good Code"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 10 - Writing good code*

----

# I: Modules, writing functions, <br> and program structure

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/16 (updated: `r Sys.Date()`)

---

## Overview of this week

.pull-left[
**This presentation - CSB 4.2:**
- Modules

- Writing your own functions

- Program structure
]

.pull-right[
**Thursday's presentation:**

- Calling Python scripts (4.4)

- Errors and error handling (4.5)

- Beyond the basics (4.9)
]

<br> <br>

.content-box-info[
**Not discussed - CSB sections on:**
- Writing style (4.3)

- Debugging (4.6), Unit testing (4.7), and Profiling (4.8)
]

---
class: center middle inverse

# Modules

-----

<br> <br> <br> <br>

---

## Modules and packages

- **Python modules** contain *collections of functions*.  
  Technically, a Python module is a single `.py` file.
  
- **Packages** are *collections of modules* packaged together   
  (though the term module is in practice also used for packages).

<br>

Use the **`import` statement** to import modules and packages into your Python
environment, be that in a script or an interactive environment:
  
  ```py
  import os
  ```

<br>

.content-box-info[
The term **library** is not as well-defined,
and can be used interchangeably with large packages,
but is also used to describe a repository of modules.
]

---

## Modules from the Python standard library

Modules from the **Python standard library**  are part of any Python
installation, but do need to be *loaded* before you can use their functionality.

For example: `csv` (working with CSV files), `os` (interfacing with the OS),
and `re` (using regular expressions &ndash; week after next).

---

## Installing modules/packages

So-called "**third-party**" modules/packages do need to be installed.

These are not installed from inside Python,
but using stand-alone package managers that can be called in the shell:

- Conda, which we have already worked with &ndash;
  for instance, to install the `numpy` ("NUMerical PYthon") library
  into an active Conda environment:
  
  ```sh
  $ conda install numpy                      # Shell
  ```

- Another common package manager is the Python-specific `pip`:
  ```sh
  $ pip install --user numpy                 # Shell
  ```

<br>

.content-box-info[
At OSC, Conda is generally the preferred method of managing Python modules.
]

---

## Installing modules/packages

.content-box-diy[

**Let's install a couple of packages,
which we'll need this week and especially next week  
into our Conda environment `ipy-env`.**

Open a terminal in VS Code, type `bash` to break out of the container, and:

```sh
sinteractive -A PAS1855 -t 60   # Start an interactive job
module load python/3.6-conda5.2 # Load the OSC Conda module
source activate ipy-env         # Activate your "ipy-env"

# Finally, install the Python packages:
conda install -y numpy pandas scipy biopython
```
]

---

## Loading modules

In Python, it is customary **not** to load modules directly into your "namespace".

This means that the module's functions will not be accessibly by just their
function names: instead, you need to **prepend the module name**.

--

- For example, simply using the module name as in `import mymodule` will load
  all of the module's functions,
  which can then be referred to using `mymodule.myfunction`:

  ```py
  import os
  os.getcwd()
  ```

--

- If a module's name is long(ish), it is common to use `as` to define an "alias"
  for the module, so the function calls can be shorter:
  
  ```py
  import numpy as np
  np.array([1, 2, 3, 4])
  ```
  
  *(While the alias name can be anything you like,
  most modules have commonly-used aliases, like `np` for `numpy`.)*

---

## Loading modules (cont.)

- To load only  certain functions from a module,
  or to load specific modules from multi-module packages:

  ```py
  # From the snakemake.io module, import two functions:
  from snakemake.io import expand, glob_wildcards
  
  # From the matplotlib package, import the pyplot module: 
  from matplotlib import pyplot as plt
  
  # Or equivalently using <package>.<module>:
  import matplotlib.pyplot as plt
  ```

<br>

--

.content-box-info[
Finally, it is *possible yet discouraged* to load all functions from a module
directly into your namespace using an asterisk `*` wildcard:
  
  ```py
  from numpy import *
  array([1, 2, 3, 4])   # Now, no need for "numpy." / "np." prefix
  ```
]

---

## Creating modules

**In Python, it is exceedingly easy to create a module:**

Just save a file with functions (e.g. `my_utils.py`) and load it using:

```python
import my_utils
```

*(This works assuming the file is in the same directory or can otherwise
be located by Python.)*

We'll do this in a little bit with our popgen simulation!

---
class: center middle inverse

# Writing your own Python functions

-----

<br> <br> <br> <br>

---

## Writing our first function

A function to calculate the **GC content of a DNA sequence** provided as a string
(the GC content is the proportion of bases that are either Gs or Cs):

```py
def GC_content(dna):

    # Make sure the input string is all uppercase:
    dna = dna.upper()
    
    # Count the occurrences of each nucleotide:
    numG = dna.count("G")
    numC = dna.count("C")
    numA = dna.count("A")
    numT = dna.count("T")
    
    # Finally, calculate (G + C) / (A + T + G + C):
    GC = (numG + numC) / (numG + numC + numT + numA)
    
    return GC 
```

.content-box-info[
Create a "cell" in your Python script using `# %%` &ndash;
now, <kbd>Shift</kbd>+<kbd>Enter</kbd> will execute the entire cell,
and create a new one below it.
]

---

## Writing our first function (cont.)

```py
def GC_content(dna):

    # Arbitrary code...
    
    return GC
```

The anatomy of a Python function:

- Start with the **keyword `def`** (for *define*)

- Next, provide the **name of the function**.
  This can be whatever we like &ndash; though it is highly recommended to be
  descriptive!
  
- Between parentheses, list the **argument(s)** (input) of the function.

- Don't forget the **colon `:`** before you start the body of the function!

- All the code belonging to the function is indented.

- The function returns output as defined after the **`return` keyword**.

---

## Using our first function

```python
GCcontent("AAATTTTG")
#> 0.125

GCcontent("ATGCATGCATGC")
#> 0.5
```

Excellent!

---

## Why are custom functions useful?

--

Custom functions are useful to break up your code in small(ish) units.

This way, the main part of your program (script) can become a series of logically
connected function calls,
and the nuts and bolts are in the individual functions.

--

<br>

**Having modular code with functions:**

- Makes it easier to see what your program does.

- Makes it easier to test your code.

- Makes it easier to repeat operations
  (like calculating GC content for multiple sequences).

- Makes your programs more flexible &ndash;
  thanks to usage of function *arguments*.

---

## Another simple function

A function to calculate the square for a series of numbers,  
and compile those squares in a list:
  
```python
def squared(start = 1, end = 10):  # We provide default values!
      
    # Create empty list to catch result of each iteration:
    results = []
      
    # Loop through the numbers:
    for i in range(start, end):
        i_square = i ** 2        # Calculate square 
        results.append(i_square) # Append current value to list
      
    return results
```

--

Let's try it:

```python
squared(start = 3, end = 10)
```

--

```python
#> [9, 16, 25, 36, 49, 64, 81]
```

---

## Arguments with default values

```python
def squared(start = 1, end = 10):
    ...
```

Because we specified default values in the function definition,  
we can omit one or both!

```python
# If we only specify "start", then "end" has default value 10:
squared(5)
#> [25, 36, 49, 64, 81]

# If we only specify "end", then "start" has default value 1:
squared(end = 3)
#> [1, 4]

# No arguments provided: start=1, end=10:
squared()
#> [1, 4, 9, 16, 25, 36, 49, 64, 81]
```

---

## Variable scope

The "scope" of a variable refers to the environment in which it can be accessed:

- When a variable has **"global scope"**, it is available anywhere in the program
  unless/until set to another value.
  This is Python's behavior for variables set anywhere *outside of functions*.

- When a variable has **"function scope"**,
  it is available only inside the function.
  This is the Python's default behavior *inside functions*.

Let's see this in action:

```python
results = [5, 10]
results
#> [5, 10]

# Our squared() function collects results in the list "results"
squared()
#> [1, 4, 9, 16, 25, 36, 49, 64, 81]

# However, after calling the function, results has been unchanged:
results
#> [5, 10]
```

---

## Variable scope

The behavior of limiting the scope of variables in functions is useful because:

--

- It means you don't have to worry about **variable names overlapping** with the
  rest of the program.

- It ensures functions don't have **"*side effects*"** apart from the single object
  they return: recall, a function exits after a statement with `return`.

*(It is possible to set variables with a global scope in functions, using the
`global` keyword, but this is not generally recommended.)*

--

.content-box-info[
So, functions are supposed to be their "own little words":

- They *should* only use objects that are explicitly passed to the
  function as arguments: **independence from the global environment**.
  
- They *should* not modify anything in the global environment except the
  single object that they return: **no side effects to the global environment**.
]

---

## <i class="fa fa-user-edit"></i> Intermezzo 4.1: <br> Determine the function of the function

```python
def foo2(x = 3, y = 5):
    if x > y:
        return x
    return y
```

--

`foo2()` returns the larger of two numbers.

---

## <i class="fa fa-user-edit"></i> Intermezzo 4.1: <br> Determine the function of the function (cont.)


```python
def foo4(x = 6):
    result = 1
    for i in range(1, x + 1):
        result = result * i
    return result
```

--

`foo4()` returns the factorial of a number (`6!` => `6 * 5 * 4 * 3 * 2 * 1`).

<br>

--

```python
def foo6(x = 25):
    if x == 1:
        return 1
    return x * foo6(x - 1)
```

--

`foo6()` returns the factorial of a number via *recursion*.

---

## <i class="fa fa-user-edit"></i> A function to convert temperatures

- Define a function that converts from temperatures in degrees Fahrenheit to
  degrees celcius.

  (To convert Fahrenheit to Celcius, subtract 32 and then multiply by 5/9.)

- Test your function with any temperature.

---

## <i class="fa fa-user-edit"></i> A function to convert temperature: Solution

```python
def convert_to_celcius(deg_f):

    deg_c = (deg_f - 32) * (5 / 9)
    
    return deg_c
```

```python
#> convert_to_celcius(100)
37.77777777777778
```

---
class: center middle inverse

# Program structure

-----

### Practicing coding a more complex program <br> with a population genetic simulation

<br> <br> <br> <br>

---

## The outline of the program

We will simulate genetic drift: random changes in allele frequency over time
that happen in finite populations.

**We'll ask how long it takes before one of the two alleles at a single locus
(`A` or `a`) drifts to fixation.**

--

<br>

To do so, we'll divide our code into several functions:

- A function that **initializes the population**.
  It takes as input the size of the population (`N`),
  and the probability of having an `A` allele (`p`),
  and returns an entire "population", i.e. a collection of diploid genotypes
  that represent individuals.

- A function that **computes genotype frequencies**,
  which we will need to determine whether an allele has gone to fixation.
  The function takes a population as input and returns the count for each
  genotype.

- A function that takes the current population and **produces the next generation**.

<br>

---

## We'll put our functions in a separate script

<br>

Start by opening a new file and saving it as `drift.py`.

---

## The function to build the population

```python
import numpy as np  # for random numbers

def build_population(N, p):
    """The population consists of N individuals. 
       Each individual has two chromosomes, containing
       allele "A" or "a", with probability p and 1-p, respectively.
       The population is a list of tuples.
    """
    
    population = []                  # Start with empty list
    
    for i in range(N):               # Loop over individuals
        
        allele1 = "A"                # The "default" allele is A
        if np.random.random() > p:   # Random number between 0 and 1
            allele1 = "a"
        
        allele2 = "A"                # Second chromosome (diploidy)
        if np.random.random() > p:
            allele2 = "a"
        
        population.append((allele1, allele2))
    
    return population
```

---

## Use Docstrings to document your functions!

.content-box-info[

Note the description of the function provided in a special multiline comment
(triple quotes), which is called a **"docstring"** (documentation string).

```python
def build_population(N, p):
    """The population consists of N individuals. 
       Each individual has two chromosomes, containing
       allele "A" or "a", with probability p and 1-p, respectively.
       The population is a list of tuples.
    """
```

<br>

We can see this documentation, for instance, if call the help for our function:

```python
>>> build_population?
>>> help(build_population)
```
]

---

## Test the function (not in the script)

With a `p` of 0.5, we expect roughly equal proportions of `A` and `a` and
differences between two instantiations:

```python
>>> build_population(N = 5, p = 0.5)
#> [('a', 'a'), ('a', 'A'), ('A', 'A'), ('A', 'a'), ('A', 'A')]

>>> build_population(N = 5, p = 0.5)
#> [('A', 'a'), ('A', 'a'), ('a', 'a'), ('A', 'A'), ('A', 'a')]
```

--

<br>

With a `p` other than 0.5, we should see uneven numbers of `a`s and `A`s:

```python
>>> build_population(N = 5, p = 0.9)
#> [('A', 'A'), ('A', 'A'), ('A', 'a'), ('A', 'A'), ('A', 'A')]

>>> build_population(N = 5, p = 0)
#> [('a', 'a'), ('a', 'a'), ('a', 'a'), ('a', 'a'), ('a', 'a')]
```

---

## The function to calculate genotype frequencies

There are four possible diploid genotypes, so our next function should simply
count the number occurrences of each genotype:

```python
def compute_frequencies(population):
    """ Count the genotypes.
        Returns a dictionary with counts for each genotype.
    """
    
    AA = population.count(("A", "A"))
    Aa = population.count(("A", "a"))
    aA = population.count(("a", "A"))
    aa = population.count(("a", "a"))
    
    return({"AA": AA,
            "aa": aa,
            "Aa": Aa,
            "aA": aA})
```

---

## Test the function (not in the script)

```python
>>> my_pop = build_population(6, 0.5)
>>> compute_frequencies(my_pop)

#>  {'AA': 1, 'aa': 1, 'Aa': 2, 'aA': 2}
```

---

## The function to create a new generation
```python
def reproduce_population(population):
    """ Create a new generation through sexual reproduction
        For each of N new offspring:
        - Choose the parents at random
        - The offspring receives one chromosome from each parent
    """
    
    new_generation = []
    
    N = len(population)
    
    for i in range(N):
        
        dad = np.random.randint(N)  # Random int between 0 & N-1
        mom = np.random.randint(N)  # (to pick indiv. from pop)
        
        chr_mom = np.random.randint(2)
        
        offspring = (population[mom][chr_mom],
                     population[dad][1 - chr_mom])
        
        new_generation.append(offspring)
    
    return(new_generation)
```

---

## The function to create a new generation
```python
def reproduce_population(population):
    """ Create a new generation through sexual reproduction
        For each of N new offspring:
        - Choose the parents at random
        - The offspring receives one chromosome from each parent
    """
    
    new_generation = []    # Start with an empty list
    
    N = len(population)    # Infer N (no need for an arg)
    
    for i in range(N):     # Run through individuals 1 at a time
        
        dad = np.random.randint(N)     # Random int between 0 & N-1
        mom = np.random.randint(N)     # (to pick indiv. from pop)
        
        chr_mom = np.random.randint(2) # Pick chromosome: 1 or 0
        
        offspring = (population[mom][chr_mom],     # 1 chr. from mom
                     population[dad][1 - chr_mom]) # 1 chr. from dad
        
        new_generation.append(offspring) # Add to master list
    
    return(new_generation)
```

---

## Test the function (not in the script)

```python
>>> reproduce_population(my_pop)
#> [('A', 'A'), ('A', 'A'), ('A', 'A'), ('A', 'a'), ('A', 'a'), ('a', 'A')]
```

---

## A new script for the main program

- Open a new file and save it as `simulate_drift.py`.

- In it, we will first import our functions:

  ```python
  import drift
  ```

<br> <br> <br>

.content-box-diy[
If your Conda installation of additional Python packages failed or is
still running, use a variant with SciPy functions instead:

```sh
# In the shell:
cd /fs/ess/PAS1855/users/$USER/week10
cp /fs/ess/PAS1855/users/jelmer/week10/drift_scipy.py drift.py
```

]
---

## Get help for our own module!

```python
>>> help(drift)
  
#> Help on module drift:
#> 
#> NAME
#>     drift
#> 
#> FUNCTIONS
#>     build_population(N, p)
#>         The population consists of N individuals. 
#>         Each individual has two chromosomes, containing
#>         allele "A" or "a", with probability p and 1-p, 
#>         respectively.
#>         
#>         The population is a list of tuples.
#>     
#>     compute_frequencies(population)
#>         Count the genotypes.
#>         Returns a dictionary with counts for each genotype.
#>     
#>     reproduce_population(population)
#>         Create a new generation through sexual reproduction
#>         For each of N new offspring:
#>         ...
```

---

## The main program, which we will also put in a function

```python
def simulate_drift(N, p):
    
    # Initialize the population:
    my_pop = drift.build_population(N, p)
    num_generations = 0   # To keep track of the generation number
    fixation = False      # Stop when fixation == True
    
    while fixation == False:
    
      # Compute genotype counts:
      genotype_counts = drift.compute_frequencies(my_pop)
      
      # If one allele went to fixation, stop:
      if genotype_counts["AA"] == N or genotype_counts["aa"] == N:
          print("An allele reached fixation at generation",
                num_generations)
          print("The genotype counts are")
          print(genotype_counts)
          
          fixation == True
          break
      
      # If not, reproduce:
      my_pop = drift.reproduce_population(my_pop)
      num_generations = num_generations + 1
```

---

## Test the program

```python
>>> simulate_drift(100, 0.5)
#> An allele reached fixation at generation 66
#> The genotype counts are
#> { ' aa ' : 100, ' aA ' : 0, ' AA ' : 0, ' Aa ' : 0}

>>> simulate_drift(100, 0.9)
#> An allele reached fixation at generation 20
#> The genotype counts are
#> { ' aa ' : 0, ' aA ' : 0, ' AA ' : 100, ' Aa ' : 0}

>>> simulate_drift(10, 0.5)
#> An allele reached fixation at generation 7
#> The genotype counts are
#> {'AA': 0, 'aa': 10, 'Aa': 0, 'aA': 0}
```

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Bonus material

-----

<br> <br> <br> <br>

---
background-color: #f2f5eb

## Saving and loading Python objects with `pickle`

It is possible to save Python objects to file "as is".

This can be handy if the object took a long time to compute,
you need to use it again in a different Python session,
and storage in a plain text file is not optimal.

`pickle` will create binary files from Python objects that can later be loaded.

- To save an object as a `pickle` file (the `.pickle` extension is not needed
  but makes it clear what the file is):

  ```python
  import pickle
  
  # "wb": Write in Binary mode
  pickle.dump(my_pop, open("population.pickle", "wb"))
  ```

- To load the object in the `pickle` file:

  ```python
  # "rb": Read in Binary mode
  population = pickle.load(open("population.pickle", "rb"))
  ```

---
background-color: #f2f5eb

## Removing some redundancy from the main function:

```python
def simulate_drift(N, p):
    
    my_pop = drift.build_population(N, p)
    num_generations = 0
    #fixation = False
    
    #while fixation == False:
    while True:
    
      genotype_counts = drift.compute_frequencies(my_pop)
      
      if genotype_counts["AA"] == N or genotype_counts["aa"] == N:
          print("An allele reached fixation at generation", num_generations)
          print("The genotype counts are")
          print(genotype_counts)
          #fixation == True
          break
      
      my_pop = drift.reproduce_population(my_pop)
      num_generations = num_generations + 1
```

---
background-color: #f2f5eb

## Another simple function

A function to nicely print a dictionary:

```python
def print_dictionary(mydic):
    
    for k, v in mydic.items():
        print("key: ", k, " value: ", v)
```

<br>

--

Let's try it:

```python
print_dictionary({"a": 3.4, "b": [1, 2, 3, 4], "c": "astring"})

#> key:  a  value:  3.4
#> key:  b  value:  [1, 2, 3, 4]
#> key:  c  value:  astring
```

---
background-color: #f2f5eb

## Keyword arguments can be passed in any order

```python
def squared(start = 1, end = 10):
    ...
```

With these so-called "keyword arguments",
we can also reverse the order of the arguments,
as long as we are calling them by name:

```python
squared(end = 5, start = 2)
#> [4, 9, 16]
squared(5, 2)
#> []
```
