---
title: "Week 10 - Writing Good Code"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 10 - Writing good code*

----

# II: Running Python scripts, errors, <br> and beyond the basics

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/18 (updated: `r Sys.Date()`)

---
class: center middle inverse

# Overview

----

### - Running Python scripts (4.4)

### - Beyond the basics (4.9)

### - Errors and error handling (4.5)

---
class: center middle inverse

# Running Python scripts (4.4)

-----

<br> <br> <br> <br>

---

## Running Python scripts

So far, we have worked with Python:
  
<<<<<<< HEAD
  - By typing in the Python interactive window or console directly.
  
  - Similarly, by interactively sending code from a script there.
  
  - By loading a Python script as a module using `import my_module`.

But, like we did with Bash scripts, we can also save a script and then  
**run it non-interactively by calling it from the shell.**

--

<br>

----

<br>

**We will practice this with our `simulate_drift.py` script from Tuesday.**

---

## Calling Python scripts (cont.)

To be able to run the script smoothly and flexibly, we will:

- Add a *shebang* line (inside the script).

- Make the script executable (outside of the script).

- Parse command-line arguments passed to the script  
  (inside & outside of the script).

- Add an `if __name__ == "__main__"` statement (inside the script).

<br>

Most of this is familiar from what we did running Bash scripts!

---

## Calling Python scripts: *shebang* line

We could write a *shebang* analogous to what we did for Bash scripts,
providing `/usr/bin/python3` (vs. `/bin/bash`) as the path to the interpreter.
This is also what CSB does.

However, we will instead **call the `env` utility to determine the path of Python
for us.**

=======
  - By typing in the Python interactive window directly.  
    And similarly, by interactively sending code from a script there.
  
  - By loading a Python script as a module using `import my_module`.

But, like we did with Bash scripts, we can also save a script and then  
**run it non-interactively by calling it from the shell.**

--

<br>

----

<br>

**We will practice this with our `simulate_drift.py` script from Tuesday.**

---

## Running Python scripts (cont.)

To be able to run the script smoothly and flexibly, we will:

- Add a *shebang* line (inside the script).

- Make the script executable (from the shell).

- Parse command-line arguments passed to the script  
  (inside the script & from the shell).

- Add an `if __name__ == "__main__"` statement (inside the script).

<br>

Most of this is familiar from what we did running Bash scripts!

---

## Running Python scripts: *shebang* line

We could write a *shebang* line just like we did for Bash scripts,
directly providing the path to the interpreter,
which in this case would be `/usr/bin/python3` (vs. `/bin/bash`).<sup>[1]</sup>

.footnote[<sup>[1]</sup>This is also what CSB does in the example.]

However, we will instead **call the `env` utility to determine the path of Python
for us.**

>>>>>>> main
--

Let's see what difference that makes &ndash; open a new terminal and type:

```sh
$ /usr/bin/env python3
#> ...
#> Python 3.6.8
>>>                     # Ctrl-D or exit() to exit Python

$ which python3
#> /usr/bin/python3
```

That loaded OSC's default Python 3 installation at `/usr/bin/python3`.

---
<<<<<<< HEAD

## Calling Python scripts: *shebang* line

=======

## Running Python scripts: *shebang* line

>>>>>>> main
If we now load our Conda environment with a new version of Python,  
and try again:

```sh
$ module load python/3.6-conda5.2
$ source activate ipy-env

$ /usr/bin/env python3
#> ...
#> Python 3.9.2
>>>                     # Ctrl-D or exit() to exit Python

$ which python3         # Check the current path to Python
#> ~/.conda/envs/ipy-env/bin/python3
```

Nice! `env` automatically tracked down the correct version of Python for us.

<<<<<<< HEAD
So, using `/usr/bin/env python3` in the *shebang* line is flexible,
and we don't need to change the path depending on which Python installation
we want to use.
=======
<br>

So, using `/usr/bin/env python3` in the *shebang* line means we don't need to
change the path depending on which Python installation we want to use.
>>>>>>> main

---

## Running Python scripts: Make the script executable

Make the script executable in the shell &ndash;
same as we've done with Bash scripts:

```sh
$ chmod u+x simulate_drift.py
```

---

## Running Python scripts: Parse command-line arguments

Again like we did with Bash scripts,
we generally want our Python scripts to take *arguments*.

Arguments that were passed to the script (as in `./my.py arg1 arg2`) are
available inside the script in the list `sys.argv`.

--

Just like `$0` contained the name of the script in Bash,
and `$1` contained the first argument, and so on:

- `sys.argv[0]` contains the name of the script.

- `sys.argv[1]` contains the first argument.

- `sys.argv[2]` contains the second argument.

- And so on...

<<<<<<< HEAD
Again like in our Bash scripts,
we also generally want scripts to take *arguments*.

Arguments that were passed to the script (as in `./my.py arg1 arg2`) are
available inside the script in the list `sys.argv`.

Just like `$0` contained the name of the script in Bash,
and `$1` contained the first argument, and so on, in Python:

- `sys.argv[0]` contains the name of the script.
- `sys.argv[1]` contains the first argument.
- `sys.argv[1]` contains the second argument.
- And so on...

---

## Calling Python scripts: Parse command-line arguments

To parse the command-line arguments, we therefore include the following code
in the script:
=======

<br>

.content-box-info[
For more sophisticated argument parsing, such as accepting flags and named options,
look into the [`argparse` module](https://docs.python.org/3/howto/argparse.html).
]

---

## Running Python scripts: Parse command-line arguments

Our `simulate_drift.py` script should take two arguments:  
the population size `N` and `p` which is essentially the starting allele
frequency.

Because the items in the `sys.argv` list are strings by default,  
we need to convert them to the right type:
>>>>>>> main

```python
import sys              # PUT THIS AT THE TOP OF THE SCRIPT

<<<<<<< HEAD
# Parse command line arguments, and convert to the right type
# (they are strings by default):
import sys ?

N = int(sys.argv[1])
=======
N = int(sys.argv[1])    # PUT THIS BELOW THE FUNCTION
>>>>>>> main
p = float(sys.argv[2])
```

<br>

<<<<<<< HEAD
Additionally, our script currently just contains a function.
We therefore also need to *call* the function in order to run it,
and we use the command-line arguments to do so:
=======
--

Additionally, our script currently just contains a *function*.  
We therefore also need to *call* the function in order to run it,  
and we use `N` and `p` as parsed from the command-line arguments to do so:
>>>>>>> main

```python
# Run the simulation:
simulate_drift(N, p)
```

---

## Running Python scripts: `__name__`

A Python script can be run in two quite different ways:

- It can be **imported into any Python session using `import`**.  
  When we do this, we typically just want to load the functions that are 
  available in the script/module to use them in our own script.

- It can be **run from the shell.**  
  In this case, we want to directly run computations,
  e.g. *call* the functions in the script. 

---

<<<<<<< HEAD
## Calling Python scripts: `__name__`
=======
## Running Python scripts: `__name__`
>>>>>>> main

We could create separate scripts for both use cases, but Python makes it easy
for **a single script to perform either of these functions depending on the context.**

Using the special variable `__name__`, we can tell the use case:

<<<<<<< HEAD
Using the special variable `__name__`, we can tell the use case:

- If the script is being imported,  
  **the value of `__name__` will be the name of the script.**

- If the script is being called from the shell,  
  **the value of `__name__` will be `'__main__'`.**
=======
| If the script is being: | Then the value of `__name__` is:
|-------------------------| -----
| Called from the shell   | `__main__`
| Imported into a Python session  | *The name of the script*

>>>>>>> main
  
---

## Running Python scripts: `__name__`

<<<<<<< HEAD
Therefore, we can include the following `if` statement to determine whether
a script is being called from the shell &mdash; and *only* if it is, we actually
parse arguments and *run* the functions in our script:
=======
Therefore, we can include an `if` statement to parse arguments and *run* the
functions in our script only when our script is being called from the shell:
>>>>>>> main

```python
if __name__ == "__main__":
    
    N = int(sys.argv[1])
    p = float(sys.argv[2])
  
    simulate_drift(N, p)
```

<<<<<<< HEAD
--

While the `__name__ == "__main__"` call is not necessary *just* for our
current use case
(we want to call the script to run it and may not have an intention to import it),
it is common and good practice for Python scripts to include this.
=======
<br>
>>>>>>> main

On the other hand, when we import the script (module),
these commands won't be executed:
its functions will just be made available to us.

---

## Finally, running our script

Now, we can run our Python script from the shell:

```sh
$ ./simulate_drift.py 1000 0.1

#> An allele reached fixation at generation 162
#> The genotype counts are
#> {'AA': 0, 'aa': 1000, 'Aa': 0, 'aA': 0}
```

---
background-color: #f2f5eb

## Side note I: <br> Standard input and standard output in scripts

Python makes it straightforward to let scripts interface with standard input
and standard output.

To **accept input from standard input**,
we can use the `sys.stdin()` function:

```python
#!/user/bin/env python
import sys

for line in sys.stdin():
    # Process each line...
```

--

To **send output to standard out**, we can simply use the `print()` function
as we did `simulate_drift.py`, or `sys.stdout()` (and `sys.stderr()`).

If our output is processed data,
printing to standard out rather than saving to file can be also useful as it
allows you to flexibly redirect output into any file,
possibly gzip it, and perhaps even pipe it into another program! 

```sh
grep "my_gene" my.GFF | ./process_gene.py | gzip > gene_output.gz
```

---
<<<<<<< HEAD

## Side note: standard in and standard out in scripts

## TODO
=======
background-color: #f2f5eb

## Side note II: <br> Using Python to orchestrate analysis pipelines

We can also run external command-line programs (and shell tools) from within
Python scripts as we can **run any run shell command** in a Python script.<sup>[1]</sup>

.footnote[<sup>[1]</sup>This can be done e.g. with the `subprocess.check_output()` function.]

Combined with Python's advanced argument parsing capabilities such as with
the `argparse` module, this means that you can **use Python to glue together
more complex analysis pipelines** than would be advisable with Bash scripts.

<br>

--

However, we will learn about an "even more advanced" of managing workflows
using Snakemake (which *is* a Python program!).
>>>>>>> main

---
class: center middle inverse

# Beyond the basics (4.9)

-----

<br> <br> <br> <br>

---

## Arithmetic of data structures

- Recall that we can use operators like `+` with lists:

  ```python
  a = [1, 2, 3]
  b = [4, 5]
  
  a + b
  ```

---

## Arithmetic of data structures

- Recall that we can use operators like `+` with lists:

  ```python
  a = [1, 2, 3]
  b = [4, 5]
  
  a + b
  #> [1, 2, 3, 4, 5]
  ```

--

- ...As well as with tuples:

  ```python
  a = (1, 2)
  b = (4, 6)
  a + b
  #> (1, 2, 4, 6)
  ```

<br>

<<<<<<< HEAD
**We can *concatenate* lists and tuples using the `+` operator.**
=======
**Specifically, the `+` operator will *concatenate* lists and tuples.**
>>>>>>> main

---

## Arithmetic of data structures (cont.)

- But we can't do the same for sets:

  ```python
  a = {1, 2}
  b = {4, 6}
  a + b
  #> TypeError: unsupported operand type(s) for +: 'set' and 'set'
  ```

- How could we combine `a` and `b` into a single set with all 4 numbers?

--

  ```python
  a | b
  {1, 2, 4, 6}
  ```

---

## Arithmetic of data structures (cont.)

- We also can't use `+` for dictionaries:
  
  ```python
  z1 = {1: "AAA", 2: "BBB"}
  z2 = {3: "CCC", 4: "DDD"}
  z1 + z2
  #> TypeError: unsupported operand type(s) for +: 'dict' and 'dict'
  ```

- How can we combine two dictionaries again?

--

  ```python
  z1.update(z2)
  z1
  #> {1: 'AAA', 2: 'BBB', 3: 'CCC', 4: 'DDD'}
  ```

---

## Arithmetic of data structures (cont.)

We can also "multiply" strings, lists, and tuples,
<<<<<<< HEAD
though recall that this repeats the contents *n* time and does not operate
=======
though recall that this repeats the contents *n* times and does not operate
>>>>>>> main
element-wise:

```python
'a' * 3
#> 'aaa'

(1, 2, 4) * 2
#> (1, 2, 4, 1, 2, 4)

[1, 3, 5] * 4
#> [1, 3, 5, 1, 3, 5, 1, 3, 5, 1, 3, 5]
```

---

## Mutable and immutable data types

*Strings*, *sets*, and *tuples* are **immutable**. As a consequence:

<<<<<<< HEAD
- They don't have methods that can modify them *in place*, so any
  "modification" in fact creates an entirely new object:
=======
- They don't have methods that can modify them *in place*,  
  so any "modification" in fact creates an entirely new object:
>>>>>>> main
  
  ```python
  seq = 'AACT'
  
  seq.replace("T", "U")
<<<<<<< HEAD
  'AACU'
  
  seq
  'AACT'
=======
  #> 'AACU'
  
  seq
  #> 'AACT'
>>>>>>> main
  ```

--

- Assigning a new variable name will always create an entirely new,
  unlinked variable:

  ```python
  x = 5
  y = x
  y = y + 10
  y
  #> 15
  x
  #> 5
  ```

---

## Mutable and immutable data types

We have seen that *lists*, *dictionaries*, and *sets* are **mutable**.
As a consequence:

- They have methods that can modify them *in place*.

  ```python
  seq = list('AACG')
  reversed_seq_maybe = seq.reverse()
  
  reversed_seq_maybe
  #>
  print(reversed_seq_maybe)
  #> None
  
  seq
  #> ['G', 'C', 'A', 'A']
  ```

---

## Mutable and immutable data types

We have seen that *lists*, *dictionaries*, and *sets* are **mutable**.
As a consequence:

- They have methods that can modify them *in place*.

- When we assign them to a new variable name, all that happens is that we
  **create a new pointer to the *same variable*,**:
  
  ```python
  seq2 = seq
  
  seq2.append('T')
  
  seq2
  #> ['A', 'A', 'C', 'G', 'T']
  seq
  #> ['A', 'A', 'C', 'G', 'T']
  ```

  To actually create a new object with the same contents,
  we need to explicitly make a copy.

---

## Copying mutable objects 

- The `copy()` method will create a copy:

  ```python
  import copy
  seq2 = seq.copy()
  seq
  #> ['A', 'A', 'C', 'G', 'T']
  ```

--

- Now, these lists can be modified independently:

  ```python
  seq2.clear()
  
  seq2
  #> []
  
  seq
  #> ['A', 'A', 'C', 'G', 'T']
  ```

--

.content-box-info[
For lists specifically, we can also use `[:]` notation:

```python
seq2 = seq[:]
```
]

---

## Copying mutable objects 

**However**, for "nested" objects, such as lists containing lists or dictionaries
containing dictionaries, the "shallow" copy made by `copy` does not suffice:

```python
a = [[1, 2], [3, 4]]
b = a.copy()

a[0][1] = 10

a
#> [[1, 10], [3, 4]]
b
#> [[1, 10], [3, 4]]     # b also changed!
```

---

## Copying mutable objects 

Instead, we need the `deepcopy()` method from the `copy` module:

```python
import copy

a = [[1, 2], [3, 4]]
b = copy.deepcopy(a)

a[0][1] = 10

a
#> [[1, 10], [3, 4]]
b
#> [[1, 2], [3, 4]]      # b has not changed
```

---
class: center middle inverse

# Errors and error handling (4.5)

-----

<br> <br> <br> <br>

---

## Errors and exceptions

Python's error messages contain a lot of useful information,  
so it really pays off to pay attention to them.

The two main error types that Python distinguishes between are:

- **Syntax errors**: You have not followed the syntax (= grammar) of Python.

- **Exceptions**: The code is gramatically correct but cannot be executed.

---

## Most common error messages

| Error type       | Explanation
|------------------|--------------
| SyntaxError      | **Code contains a syntax (i.e., grammar) error**. <br> Common problems are a missing colon, missing quotes, `=` instead of `==`.
| IndentationError | **Incorrect amount of indentation**, <br> e.g. after `if`, loop, or function definition.
| TypeError        | **Variable type does not allow the operation**: <br> e.g. adding a number and a string.
| NameError        | **Variable does not exist** (typo?).
<<<<<<< HEAD
=======

---

## Most common error messages

| Error type       | Explanation
|------------------|--------------
| SyntaxError      | **Code contains a syntax (i.e., grammar) error**. <br> Common problems are a missing colon, missing quotes, `=` instead of `==`.
| IndentationError | **Incorrect amount of indentation**, <br> e.g. after `if`, loop, or function definition.
| TypeError        | **Variable type does not allow the operation**: <br> e.g. adding a number and a string.
| NameError        | **Variable does not exist** (typo?).
>>>>>>> main
| IndexError       | **Item in list does not exist.**
| KeyError         | **Item in dictionary does not exist.**
| IOError          | **File** does not exist, or you tried to write to a file that is open only for reading (IO = Input/Output).
| AttributeError   | **Attribute or method does not exist**, <br> e.g. `.keys()` on a list.

---

## Handling exceptions

<<<<<<< HEAD
Like almost any programming language (but unlike Bash),
Python will stop execution when it encounters an error.

So, when an error is encountered in the middle of running your script,
=======
Like most programming languages (but unlike Bash),  
Python will **stop execution** when it encounters an error.

So, when an error is encountered in the middle of running your script or function,
>>>>>>> main
any code after the error will not be run.

For instance, we may get an error when trying to divide by 0:

```python
def div(x, y):
    print("I will divide!")
    print(x / y)
    print("I'm done")
    
div(10, 5)
#> I will divide!
#> 2
#> Still here?

div(10, 0)
#> I will divide!
#> [...]
#> ZeroDivisionError: float division by zero
```

---

## Handling exceptions (cont.)

Sometimes, you want to allow for an error and carry on,  
in which case you are said to **"catch" the error**.

For example:

```python
def superdiv(x, y):
    print("I will divide!")
    try:
        print(x / y)
    except:
        print("Error: Cannot divide by 0")
    print("I'm done")


superdiv(10, 0)
#> I will divide!
#> Error: Cannot divide by 0
#> I'm done
```

<<<<<<< HEAD
## expand
<br>
=======
---

## Handling exceptions (cont.)

A less silly version of that,
where you actually need to accommodate zero-divisions,
could look something like this:

```python
def superdiv(x, y):
    try:
        return(x / y)
    except:
        return(None)      # Return None when trying to divide by 0
```
>>>>>>> main

--

<br>

.content-box-info[
Note that this is doing the opposite of what we were doing in Bash scripts,
where we made sure the program stops when it encounters an error,
which it did not do by default.

The distinction is that in *specific, anticipated cases*,
we actually may want to carry on despite something that is technically an error.
]

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Bonus material

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Writing style (4.3)

-----

<br> <br> <br> <br>

---
background-color: #f2f5eb

## Some style  guidelines

- Use four spaces per indentation level; not tabs.

- Split long lines using parentheses: consider having your text editor insert
  a vertical line at 80 characters and try not to pass that line.

- Separate functions using blank lines.

- Use a new line for each operation:

.pull-left[
Yes:
```python
import scipy
import re
```
]

.pull-right[
No:
```python
import scipy, re
```
]

<br> <br>

.content-box-info[
For more style guidelines, see
[PEP8 - The style guide for Python code](https://pep8.org/)
]


---
background-color: #f2f5eb

## Some style guidelines (cont.)

- Put spaces around operators (unless using the : to “slice” a list).

- Use parentheses to clarify the precedence of operators.

.pull-left[
Yes:
```python
a = b ** c
z = [a, b, c, d]
n = n * (n - 1)
k = (a * b) + (c * d)
my_list = a_list[1:n]
```
]

.pull-right[
No:
```python
a=b**c
z=[ a,b,c,d ]
n =n*(n-1)
k=a*b+c*d
my_list = a_list[1 : n]
```
]

---
background-color: #f2f5eb

## Some style guidelines (cont.)

- **Use lots of comments and "docstrings"!**

  Python allows you to write short comments (starting with # ) and
  docstrings (multiline comments flanked by triple double quotes, """ ):
  
  - Use docstrings to document *how to use the code*.  
    What does this function do? What is the input/output?
  
  - Use short comments for explaining difficult passages in the code:  
    that is, to document how the code works.


<br>

--

- **Naming conventions:**

  - Variables and functions in lowercase, "snake case": `my_var`, `my_fun`.
  
  - **Constants** in uppercase: `SOME_CONSTANT`.
  
  - Use descriptive names: `count_pairs`, `body_mass`, `cell_volume`.
  
  - Verbs for functions and nouns for objects.
