---
title: "Week 9 - Control flow and working with files - II"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 9 - Control flow and working with files*

----

# II: Working with files

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/11 (updated: `r Sys.Date()`)

---

## Getting set up

1. In the shell, look for your `CSB` dir (the book's GitHub repo) or
   download it again:

   ```sh
   git clone https://github.com/CSB-book/CSB.git         # SHELL
   ```

2. Trigger an interactive Python window from a Python script
   (<kbd>Shift</kbd>+<kbd>Enter</kbd>).

3. Move to the `python/sandbox` dir in CSB in Python.

   Assuming your CSB dir is at `/fs/ess/PAS1855/users/$USER/CSB`,  
   and your Python script in `/fs/ess/PAS1855/users/$USER/week09`: 

  ```py
  import os                                             # PYTHON
  os.chdir('../CSB/python/sandbox')
  
  ## Or w/ an absolute dir, replace <username> w/ real username:
  # os.chdir('/fs/ess/PAS1855/users/<username>/CSB/python/sandbox')
  
  ## Check your current working dir:
  # os.getcwd()
  ```

---

## CSB 3.7: Working with files

In Python, we need to *explicitly open a file* before we can interact with it.
Opening a file won't load it into memory, but will create a connection to the file.

For instance, to open a file for *writing* (`w`):

```python
my_filehandle = open('my_new_file.txt', 'w')
```

Now, `my_filehandle` is is a file object, also called a *file handle*.

<br>

<figure>
<p align="center">
<img src=img/file_handle_output-3.png width="65%">
<figcaption>From O’Neil 2017 - A Primer for Computational Biology</figcaption>
</p>
</figure>

---

## Most commonly used file modes

.center[
| Shorthand | File mode
|-----------|----------
| `r`       | Read
| `w`       | Write
| `a`       | Append
| `r+`      | Read + write
]

<br>

```python
open('my_new_file.txt', 'w')
open('some_data.txt', 'r')
```


---

## Writing to a file

To actually write to a file, we can use the `write()` method to write a string
to the file, or `writelines()` to write a list to the file.

- However, in either case, we need to insert the newlines ourselves with `\n`:  

  ```python
  s = 'My file contents'
  my_filehandle.write(s + "\n")
  
  my_filehandle.writelines(["A\n", "B\n", "C\n"])
  ```

--

- When we are done with the file, **we need to close it**:

  ```python
  my_filehandle.close()
  ```

<br>

.content-box-info[
To write a list to file one line at a time, use `join()`:
```py
f.write("\n".join(mylist))
```
]

---

## A more convenient way: `with open()`

- If we use the `with open()` construct,
  we don't have to manually close the file &ndash;
  instead, the file connection is open as long as our code is *indented*:
  
  ```python
  with open("my_new_file.txt", "r") as my_filehandle:    
    
        for my_line in my_filehandle:  # Loop over lines
            print(my_line)             # Print line back
  
  #> My file contents
  #>
  #> A
  #>
  #> B
  #>
  #> C
  ```

- The above example also shows that **if we loop over the file handle,
  we automatically process one line at a time.**
  
- Here, we simply printed the contents of the file to screen,
  but notice how we ended up with *double-spacing*.

---

## More newlines

- To see what's going on with the double-spacing,
  let's read a single line using `readline()` and store the result in a variable:
  
  ```python
  with open('my_new_file.txt', 'r') as my_filehandle:
        line1 = my_filehandle.readline()
  
  line1
  #> 'My file contents\n'
  print(line1)
  #> My file contents
  #>
  ```

--

- As it turns out, reading methods also return the newline characters `\n`.

- `print()` parses the newline character and because it by default always inserts
  a newline anyway, we end up with double-spacing.

- For single-space `print()`ing, or when processing a line,
  we can use `strip()` to remove newlines characters (and other whitespace):
  
  ```py
  line1 = my_filehandle.readline().strip()
  ```

---

## Progressively reading a file

- When reading a single line at a time using `readline()`,  
  **reading is done progressively**, so if we call `readline()` twice,
  we will read the first and *then* the second line:
  
  ```python
  with open('../data/Singh2015_data.csv', 'r') as input_file:
      
        line1 = input_file.readline().strip()
        line2 = input_file.readline().strip()
        print(line1)
        print(line2)
  
  #> Line,InfectionStatus,RecombinantFraction
  #> 21,I,0.1826923077
  ```

--

- Or to skip a line, for instance a header, we can use `next()`:

  ```python
  input_file.next()
  ```
  
---

## Summary of basic options to read from a file

- Note that with all approaches,
  line breaks are present in the resulting strings as `\n`.  

| Approach                   | Result           |
|----------------------------|------------------|
| **`for <line> in <fhandle>:`** | Loop over lines one at a time.
| **`readline()`** | Return current line as a string.
| **`readlines()`** | Return a list of strings, one item per line.
| **`read()`** | Return entire file content as a single string.

---

## Character-delimited (tabular) files

We start by reading the first few lines of a CSV (comma-separated values)
file:

```python
with open("../data/Dalziel2016_data.csv") as dalziel_file:
    
    for i, line in enumerate(dalziel_file):
        
        print(line.strip())
        
        if i > 2:
            break

#> biweek,year,loc,cases,pop
#> 1,1906,BALTIMORE,NA,526822.1365
#> 2,1906,BALTIMORE,NA,526995.246
#> 3,1906,BALTIMORE,NA,527170.1981
```

- We didn't specify an option to `open()`, so it defaults to *read* mode.

- With `enumerate()`, we can track line numbers along with line contents.

- We break out of the loop, i.e. we stop it from running,
  after we've printed the line with a line index `i` of 3 &ndash;
  the effect is that we only print the first four lines.

---

## Using the `csv` module

While we could manually split rows by commas,
a "module" that can do this for us is `csv`,
which we load using an `import` statement (more next week): 

```python
import csv

with open("../data/Dalziel2016_data.csv") as dalziel_file:
    
    # We use the function "reader" from the csv module;
    # Note the csv.reader = <module>.<function> syntax
    reader = csv.reader(dalziel_file)
    
    for i, row in enumerate(reader):
        print(row)
        if i > 2:
            break

#> ['biweek', 'year', 'loc', 'cases', 'pop']
#> ['1', '1906', 'BALTIMORE', 'NA', '526822.1365']
#> ['2', '1906', 'BALTIMORE', 'NA', '526995.246']
#> ['3', '1906', 'BALTIMORE', 'NA', '527170.1981']
```

The output is a list for every row, with one item per field.

---

## Using the `csv` module (cont.)

The `csv.DictReader()` function will return a *dictionary* for every row,
containing the *column names as the keys* and the cell values as the values:

```python
with open("../data/Dalziel2016_data.csv") as dalziel_file:
    
    reader = csv.DictReader(dalziel_file)
    
    for i, row in enumerate(reader):
        print(dict(row))
        
        if i > 2:
            break
    
#> {'biweek': '1', 'year': '1906', 'loc': 'BALTIMORE', 'cases': 'NA', 'pop': '526822.1365'}
#> {'biweek': '2', 'year': '1906', 'loc': 'BALTIMORE', 'cases': 'NA', 'pop': '526995.246'}
#> {'biweek': '3', 'year': '1906', 'loc': 'BALTIMORE', 'cases': 'NA', 'pop': '527170.1981'}
#> {'biweek': '4', 'year': '1906', 'loc': 'BALTIMORE', 'cases': 'NA', 'pop': '527347.0136'}
```

---

## Using the `csv` module (cont.)

Now, we can easily select rows based on their values in a specific column.

For instance, to take only rows from Washington &ndash;
i.e. where the column `loc` has the value "Washington":

```sh
with open("../data/Dalziel2016_data.csv") as infile:
    
    reader = csv.DictReader(infile)

    for row in reader:
        if row['loc'] == "WASHINGTON":
            print(row)
            
#> {'biweek': '14', 'year': '1940', 'loc': 'WASHINGTON', 'cases': '3', 'pop': '672643.7152'}
#> {'biweek': '15', 'year': '1940', 'loc': 'WASHINGTON', 'cases': '3', 'pop': '673370.2202'}
#> {'biweek': '16', 'year': '1940', 'loc': 'WASHINGTON', 'cases': '3', 'pop': '674095.5105'}         
#> ...
```

---

## Using the `csv` module (cont.)

Next, let's write the rows for Washington to a new, tab-delimited file:

```python
with open("../data/Dalziel2016_data.csv") as infile:
    
    reader = csv.DictReader(infile)
    
    with open('Dalziel_Washington.csv', "w") as outfile:
        
        writer = csv.DictWriter(outfile,
                                fieldnames = reader.fieldnames,
                                delimiter = '\t')
        
        writer.writeheader()   # Write the header row

        for row in reader:
            if row['loc'] == "WASHINGTON":
                writer.writerow(row)
```

- We can specify the delimiter to be a tab `\t` (default is a comma).

- `writeheader()` and `writerow()` will write data from the file *as is*.

---

## <i class="fa fa-user-edit"></i> Intermezzo 3.4

Write code that prints (using the `print()` function) the values in the columns
`loc` and `pop` for all rows in the file `Dalziel2016_data.csv`.

<!-- `r icon::fa("user-edit")` -->

---

## <i class="fa fa-user-edit"></i> Intermezzo 3.4: solutions

```py
import csv

with open ("../data/Dalziel2016_data.csv") as fhandle:
  
    reader = csv.DictReader(fhandle)
  
    for row in reader:
        print(row["loc"], row["pop"])
```

--

<br>

.content-box-info[
Or to write this to a file:

```py
with open("../data/Dalziel2016_data.csv") as infile:

    reader = csv.DictReader(infile)

    with open('Dalziel_loc-pop.csv', "w") as outfile:

        for row in reader:
            outfile.write(str(row['loc'] + ',' + row['pop'] + '\n'))
```
]

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>


---
background-color: #f2f5eb

## Reading and writing

- As another toy example, we can write the lines that we read from one file
  to another file:

  ```python
  with open('../data/Singh2015_data.csv', 'r') as input_file:
      
      with open('Singh_copy.csv', 'w') as output_file:
        
        for line in input_file:
            output_file.write(line)
  ```

- Now, we don't need to think about the `\n` characters:
  recall that reading methods *include* them,
  while writing methods "*need*" them.
  
<br>

----
  
- When reading a file line-by-line, to go back to a specific line
  in this case the 1st line, we can use:

  ```python
  input_file.seek(0)
  ```
