---
title: "Week 10 - Writing Good Code"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 10 - Writing good code*

----

# II

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/18 (updated: `r Sys.Date()`)

---
class: center middle inverse

# Overview
----

### - Writing style (4.3)

### - Calling Python scripts (4.4)

### - Errors and error handling (4.5)

### - Beyond the basics (4.9)

---
class: center middle inverse

# Writing style (4.3)

-----

<br> <br> <br> <br>

---

## Some style  guidelines

LINK TO STYLE GUIDELINES

- Use four spaces per indentation level; not tabs.

- Split long lines (especially the arguments of functions) using parentheses.

- Separate functions using blank lines.

- Use a new line for each operation:

.pull-left[
Yes:
```python
import scipy
import re
```
]

.pull-left[
No:
```python
import scipy,re
```
]

---

## Some style guidelines (cont.)

- Put spaces around operators (unless using the : to “slice” a list); use
parentheses to clarify the precedence of operators:

.pull-left[
Yes:
```python
a = b ** c
z = [a, b, c, d]
n = n * (n - 1)
k = (a * b) + (c * d)
my_list = a_list[1:n]
```
]

.pull-right[
No:
```python
a=b**c
z=[ a,b,c,d ]
n =n*(n-1)
k=a*b+c*d
my_list = a_list[1 : n]
```
]

---

## Some style guidelines (cont.)

- **Use lots of comments and "docstrings"!**

  Python allows you to write short comments (starting with # ) and
  docstrings (multiline comments flanked by triple double quotes, """ ):
  
  - Use docstrings to document *how to use the code*.
    What does this function do? What is the input/output?
  
  - Use short comments for explaining difficult passages in the code,
    that is, to document how the code works.


<br>

- **Naming conventions:**

  - Variables and functions in lowercase, "snake case": `my_var`, `my_fun`.
  
  - **Constants** in uppercase: `SOME_CONSTANT`.
  
  - Use descriptive names: `count_pairs`, `body_mass`, `cell_volume`.
  
  - Verbs for functions and nouns for objects.

---

.panelset[
.panel[.panel-name[R Code]

```{r panel-chunk, fig.show='hide'}
# ... r code ...
```
]

.panel[.panel-name[Plot]

My plot
]
]

---
class: center middle inverse

# Calling Python scripts (4.4)

-----

<br> <br> <br> <br>

---

...

---
class: center middle inverse

# Errors and error handling (4.5)

-----

<br> <br> <br> <br>

---

...

---
class: center middle inverse

# Beyond the basics (4.9)

-----

<br> <br> <br> <br>

---

## Modifying lists: in-place-acting methods

We've seen that many of the methods in the previous slides modify lists
*in place*, which is something you may not be used to.

Such methods, therefore, don't return anything:

```python
reversed_list_maybe = seq.reverse()

reversed_list_maybe
#>

seq
#> ['A', 'A', 'F', 'K', 'N', 'T', 'V', 'V']
```

---

## Modifying lists: >1 name referring to the same list

Also of note is that when we assign an existing list to a new name,  
this will *not* create a new list in the way we saw with simple variables.

```python
list_b = my_list
```

Instead, both names (`list_b` and `a`) now are separate pointers to the *same*
list &ndash; if we modify one, the other will also change:

```python
list_b = list_b.append("Exclusive to list b?")

my_list
#> [1, 2, 5, 14, 42, 132, "Exclusive to list b?"]
```

--

<br>

.content-box-info[
Python exhibits this type of behavior for all mutable data types.

To break this "linkage", we need to explicitly create a copy of the list
(next slide).
]

---

## Strings versus lists

- Because strings are immutable, assigning a new variable name will always create
  an entirely new, unlinked variable &ndash; recall from Tuesday:

  ```python
  x = 5
  y = x
  y = y + 10
  y
  #> 15
  x
  #> 5
  ```

--

- We saw earlier that strings can be indexed like lists:

  ```python
  seq = "ATGCATG"
  seq[2]
  #> 'T'
  ```

--

- So can we use this for assignment too?
  No, because strings are immutable unlike lists.

  ```python
  seq[2] = "U"
  #> TypeError: 'str' object does not support item assignment
  ```

---

## Copying lists

- `copy` &ndash; Create a copy of the list:

  ```python
  import copy
  new_list = my_list.copy()
  new_list
  #> ['blue', 2.44, 'green', True, 25, 'red', 0] 
  ```

- Or use `[:]` notation:

  ```python
  another_list = my_list[:] 
  ```

--

- Now, these lists can be modified independently:

  ```python
  new_list.clear()
  my_list
  #> [1, 2, 5, 14, 42, 132]
  
  another_list.clear()
  my_list
  #> [1, 2, 5, 14, 42, 132]
  ```

