---
  title: "Week 13 - Reproducible workflows with Snakemake"
output:
  xaringan::moon_reader:
  seal: false
css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
lib_dir: libs
nature:
  highlightStyle: rainbow
highlightLines: true
countIncrementalSlides: false
---
  class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 13 - Reproducible workflows with Snakemake*

----

# I: ...

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/23 (updated: `r Sys.Date()`)

---

## Overview of this week

---

## The need for "formal" workflow management

> The essence of what we're trying to do:
>
> - Automate a workflow
> - Document the workflow
> - Document the dependencies among data files, code
> - Re-run only the necessary code, based on what has changed
> [Karl Broman](http://kbroman.org/Tools4RR/assets/lectures/01_intro.pdf)

---

## The need for "formal" workflow management

<p align="center">
<img src=nature-feature.png width=100%>
</p>


> Typically, researchers codify workflows using general scripting languages such as Python or Bash. But these often lack the necessary flexibility. Workflows can involve hundreds to thousands of data files; a pipeline must be able to monitor their progress and exit gracefully if any step fails. And pipelines must be smart enough to work out which tasks need to be re-executed and which do not.

---

## The need for "formal" workflow management

<p align="center">
<img src=img/nature-feature.png width=100%>
</p>

> Bioinformatician Davis McCarthy at St Vincent’s Institute of Medical Research in Fitzroy, Australia, says Python and R were more than enough for the relatively simple workflows he used as a PhD student.

---

<p align="center">
<img src=img/giaa140fig1.jpeg width=100%>
</p>
