---
title: "Week 13 - Reproducible workflows with Snakemake"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
#xaringanExtra::use_scribble()
#xaringanExtra::use_panelset()
#xaringanExtra::use_clipboard()
#xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 13 &ndash; <br> Reproducible workflows with Snakemake*

----

# III: More advanced Snakemake tips and tricks

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/04/08 (updated: `r Sys.Date()`)

---
class:inverse middle center

# Overview

----

.left[
- #### Using Conda environments
- #### Useful command-line options
- #### What will Snakemake run? &ndash; revisited
- #### Configuration files
- #### Referring to wildcards in actions
- #### Miscellaneous
]

---
class: middle inverse center

# Using Conda environments

----

<br> <br> <br> <br>

---
## Using Conda environments

Snakemake plays well with Conda and with Singularity containers,
and using such software environments can help a lot with making your workflow
functional and reproducible across different hardware environments.

While containers are even better at this, Conda is more lightweight
and will suffice for most purposes.

Snakemake will in fact **create the necessary Conda environments for us**,  
all we need to do is:

- Include a `conda` key in the focal rule,
  where we specify a YAML file with instructions for the environment:

  ```python
  rule fastqc:
      conda: "envs/fastqc-env.yaml"
      # ...other keys...
  ```

- Include the `--use-conda` option in our Snakemake call:

  ```sh
  snakemake -j1 --use-conda
  ```

---

## Using Conda environments (cont.)

Recall that we can create such YAML files from existing Conda environments
using `conda env export`,
but it is also very easy to write a file from scratch.

For instance, for a specific version of FastQC, and specifying `bioconda` as
the channel, the full YAML file would look like this:<sup>[1]</sup>

```sh
name: fastqc-env
channels:
  - bioconda
dependencies:
  - fastqc=0.11.8
```

.footnote[
<sup>[1]</sup> For use with Snakemake, the `name` key can also be omitted.  
]

---
class: inverse middle center

# Useful command-line options

----

<br> <br> <br> <br>

---

## Useful command-line options: dry run (-n)

With the `-n` (long form: `--dryrun`) option, Snakemake will not run any of
the actions specified, but it will check which rules/jobs it *would* be running.

--

<br>

This can be extremely useful to:
  
- Check whether the workflow will run the way you think it will run
  (remember: relationships between rules are not explicitly defined...).
    
- Check if anything needs to be done: you may not remember which output files
  are already there, or if an input file has been updated, and this is the
  easiest way to check!
    
- Check if you have no errors in your Snakefile
  (e.g. syntax errors, wildcards that Snakemake can't figure out).

<br>

```sh
snakemake -j1 -n
```

---

## Useful command-line options: <br> reason (-r) and print shell commands (-p)

We have already seen these two options:

- `-r` (long form: `--reason`) to let Snakemake tell you *why* it thinks what it
  should run.

- `-p` (long form: `--printshellcmds`) to print the shell commands
  (action key `shell`) that are being run.

<br>

*These are very useful to also use in combination with the `-n`/`--dryrun` option,
for troubleshooting, and for getting a better intuition for how Snakemake works.*

<br>

```sh
snakemake -j1 -npr
```

---

## Useful command-line options: Miscellaneous

Short  |  Long            | Explanation
-------|------------------|-------------
-      | `--lint`            | Run the Snakemake "linter" on the Snakefile
`-f`   | `--force <target>`  | Force creation of a specified target file, or the first rule.
`-F`   | `--forceall <rule>` | Force running a specified rule and all dependencies.
`-R`   | `--forcerun`        | Force creation of a list of target files.
-      | `--report`          | Create an HTML report with runtime statistics, workflow topology, and optionally results ([documentation](https://snakemake.readthedocs.io/en/stable/snakefiles/reporting.html)).
-      | `--archive` <span style="color:#e8e8e8">`_________________`</span> | Create a self-contained workflow archive that could be re-executed on any machine with Snakemake and Conda.  

---

## Useful command-line options: Miscellaneous

**Overview of options we have seen so far:**

Short  |  Long            | Explanation
-------|------------------|-------------
`-q`   | `--quiet`        | Less output, can be useful with `-n` just to get overview of jobs that will be run.
`-p`   | `--printshellcmds` | Print `shell` commands that will be executed.
`-r`   | `--reason`       | Give reason of execution for every job.
`-n`   | `--dryrun`       | Don't run anything, just report what *would* be run.
`-j`   | `--job`          | Max. number of jobs to run simultaneously.
`-s`   | `--snakefile` <span style="color:#e8e8e8">`_________________`</span> | Name of / path to the Snakefile.

---
class: middle inverse center

# What will Snakemake run? &ndash; Revisited

----

<br> <br> <br> <br>

---

## What will Snakemake run?

- **Snakemake will create an output file, and all its dependencies, if:**
  
  - The output file is a command-line target and does not exist.
  
  - The output file is needed by another executed job and does not exist.
  
  - The output file is older than the input file.
  
  - The input file will be updated by other job.
  
  - Execution is enforced with `-f` / `-F` / `-R`.

---

## What will Snakemake run? (cont.)

- **Snakemake will not create output files *even if*:**

  - Code in the rule (or elsewhere in the Snakefile) has changed.
  
  - Additional input files have been added *and* input files are e.g. listed
    at the top of the file or in a config file.
    (With `glob_wildcards()`, new files will be automatically run.)

  *It's possible to let Snakemake detect such changes, though, and then force
  a rerun for those &ndash;
  see [here](https://snakemake.readthedocs.io/en/stable/project_info/faq.html#snakemake-does-not-trigger-re-runs-if-i-add-additional-input-files-what-can-i-do)
  and the exercises for more details.*

---
class: middle inverse center

# Configuration files

----

<br> <br> <br> <br>

---

## Configuration files

*Configuration files* can be a good place to store settings for your workflow
that you want to be able to vary easily and transparently,
or that you just want to state very clearly.

In your Snakefile, you can include a `configfile` directive
(for the entire workflow, not for individual rules) that points to a YAML or
JSON file with these configurations.

<br>

--

Then, all key-value pairs from the config file will loaded into a dictionary
called `config` that you can call in your Snakefile &ndash; for example:

```sh
# CONFIG FILE:
min_qual: 30
```

```python
# SNAKEFILE:
config["min_qual"]   # Get the value for `min_qual`
```

---

## Example usage of a configuration file

A config file called `config.yaml` contains the following:

```sh
out_dir: path/to/output/
min_qual: 30
```

In the Snakefile, we access the values in the config files as follows:

```python
# [...]
configfile: "config.yaml"
OUT_DIR=config["output_dir"]

# [...]

rule filter_bam:
    input: ...
    output: os.path.join(OUT_DIR, "{sample}_{read}.fastqc.html")
    shell: 'filter_bam.sh -q config["min_qual"] {input} > {output}'
```

Note that we assigned `output_dir` to a variable in the Snakefile,
and used `min_qual` directly &ndash; either is possible.

---
class: middle inverse center

# Referring to wildcards in actions

----

<br> <br> <br> <br>

---

## Referring to wildcards in actions

Sometimes you need to **refer to a sample ID in your action** rather than an
input file name &ndash;
say our script needs an input dir and sample/book ID like so:

```sh
scipts/count_words.py -d <input_dir> -i <book_id>
```

Our rule currently looks like this:

```python
rule count_words:
    input: "data/{book}.txt",
    output: "res/{book}_cnt.txt"
    shell: "scripts/count_words.py {input} > {output}"
```

**How can we change the `shell` directive to use the book ID wilcard?**

---

## Referring to wildcards in actions (cont.)

**How can we change the `shell` directive to use the book ID wilcard?**

```python
# THIS WON'T WORK:
# shell: "scripts/count_words.py -d res -o {book} > {output}"
```

```python
# YES -- use "{wildcards.<wildcard-name>}":
shell: "scripts/count_words.py -d res -o {wildcards.book} > {output}"
```

Why do we need `{wildcards.book}` and not just `{book}` akin to how we refer to
`{input}`?

--

Even though both are variable-like references with `{}`,
only `{book}` is a **wildcard**,
which has not been explicitly defined, and is left for Snakemake to figure out.

Snakemake figures this out in `{input}` and `{output}` directives *only*,
and if we want to refer to them elsewhere,
we need to use `{wildcards.<wildcard-name>}`.

---
class: middle inverse center

# Miscellaneous

----

<br> <br> <br> <br>

---

## Miscellaneous

### The params directive

### TODO

<br>

### Call separate Snakefiles for parts of the workflow

If your workflow is pretty big, it can be useful to split it up across multiple
Snakefiles. You can make rules in another Snakefile available as follows:

```python
include rules/qc_rules.smk
```

---

## Miscellaneous (cont.)

### Mark files as protected (no write permissions)

```python
output: protected("sorted_reads/{sample}.bam") 
```

<br>

### Mark files as temporary (to be deleted)

```python
output: temp("mapped/{sample}.bam")
```

---

## Miscellaneous (cont.)

### Use a "token" file if a rule has no unique output

(e.g. only modifies a file) -- and `touch` in the rule's action:

```python
rule token_example:
    input:  'some_file.txt'
    output: 'some_file.tkn'
    shell: "some_command --do-things {input} && touch {output}"
```

<br>

### Visualize rules, not jobs

```python
snakemake --rulegraph | dot -Tsvg > rules.svg
```

---

## Miscellaneous (cont.)

### Calling Python or R scripts directly, without arguments

- When calling scripts, "snakemake.input" etc is available automatically:

```python
script: pythonscript.py
    quals = [rec.qual for record in VarFile(snakemake.input[0])]
script: rscript.R
    snakemake@input[["myfile"]]                 # In R, S4 object will be available
```
