---
title: "Week 11 - Python: Scientific computing"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Scientific computing*

----

# I: NumPy

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/23 (updated: `r Sys.Date()`)

---

## Overview of this week

- **NumPy** (CSB 6.2) &ndash; Tuesday

- **Pandas** (CSB 6.3) &ndash; Tuesday/Thursday

- **BioPython** (CSB 6.4) &ndash; Thursday

---
class: center middle inverse

# NumPy (CSB 6.2)

-----

<br> <br> <br> <br>

---

## What is NumPy?

NumPy is the basic Python package for scientific computing.

- NumPy makes working with **numeric data** in Python a lot easier.

- It has an efficient data structure, the **NumPy array**,
  for vectors, matrices, and multidimensional arrays.

- These NumPy arrays allow for **vectorized computation**.

- Some other functionality includes the ability to **generate random numbers**,
  for which we have already used NumPy a bit last week.

--

<br>

As we have seen, NumPy is typically imported with `as np` added,  
so we can access its functions with a little less typing:

```python
import numpy as np

# Now we access numpy functions with "np.<function-name>"
```

---

## NumPy arrays

**Arrays** are NumPy's main data structure, which can contain:

- Numerical *vectors* &ndash; 1-dimensional arrays
- *Matrices* &ndash; 2-dimensional arrays
- *Tensors* &ndash; n-dimensional arrays

--

<br>

For example, to create a **one-dimensional array**,  
we can use NumPy's variant on the `range()` function, `np.arange()`:

```python
a = np.arange(9)
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```

<br>

One of the key differences between regular Python lists and NumPy arrays is
that **all elements in an array have to be of the same type.**  
(And usually, but not necessarily, they will be integers or floats.)

---

## NumPy arrays (cont.)

- Recall once again that with lists,
  we can use operators like `*` but the operations are not done element-wise:

  ```python
  mylist = list(range(9))
  mylist * 2
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8, 0, 1, 2, 3, 4, 5, 6, 7, 8]
  ```

--

- However, with arrays, these operators do work **element-wise**:

  ```python
  # Multiply each element by 2:
  a * 2
  #> array([  0,  2,  4,  6,  8, 10, 12, 14, 16])
  
  # Multiply values element-by-element:
  a * a
  #> array([ 0,  1,  4,  9, 16, 25, 36, 49, 64])
  ```
  
  Such operations are said to be "**vectorized**".  
  
  In the first example, `2` is recycled (or **_broadcast_**) as many times as
  is necessary to multiply each element in `a` by `2`.

---

## Some attributes of NumPy arrays

- Get the length (number of elements) along each dimension with `shape`:

  ```python
  a.shape   # This is an attribute, not a method - no parentheses!
  #> (9,)       
  ```

  The result is returned as a *tuple*, in this case containing only one element
  (note the trailing comma), since we only have one dimension.
  
--

- Get the number of dimensions with `ndim`:

  ```python
  a.ndim
  #> 1
  ```

- Get the number of elements with `size`:
  
  ```python
  a.size
  #> 9
  ```

---

## Some attributes of NumPy arrays (cont.)

Get the data type for the elements in the array with **`dtype`**:
  
```python
a.dtype
#> dtype('int64')        # 64-bit integers
```

Because all elements in the array have to be of the same type,
the type information is stored at the array-level and not at the element-level.

<br>

.content-box-info[
**Bits in data types**  
Above, the data type was reported to be **`int64`** rather than just `int`.
NumPy has separate data types for numbers stored with different numbers of bits,
so there is `int8`, `int16`, `int32`, `int64`,
and even the same sequence for "unsigned integers", integers that are positive
only (e.g. `uint64`).

*For the most part, you don't have to pay attention to this* &ndash;  
and when assigning data types, you can just use `int` or `float` without
specifying bits.
]

---

## Simple arithmetic and statistics for arrays

- Sum of all elements with `sum()`:

  ```python
  a.sum()
  #> 36
  ```
  
- Mean and standard deviation with `mean()` and `std()`:
  
  ```python
  a.mean()
  #> 4.0
  
  a.std()
  #> 2.5819888974716112
  ```

--

- Lowest and highest value with `min()` and `max()`:

  ```python
  a.min()
  #> 0
  
  a.max()
  #> 8
  ```

---

## Creating arrays &ndash; by converting lists

**Converting a list to a one-dimensional array**

- If we create an array of numbers without decimal points,  
  they will be interpreted as integers:

  ```python
  a1 = np.array([1, 2, 3, 4])
  print(a1)
  #> [1 2 3 4]
  
  a1.dtype.name
  #> 'int64'
  ```

--

- If we include decimal points, they will be floats:

  ```python
  a1 = np.array([1.0, 2.0, 3.0, 4.0])
  
  print(a1)
  #> [ 1. 2. 3. 4.]
  a1.dtype.name
  #> 'float64'
  ```

---

## Creating arrays &ndash; by converting lists (cont.)

**"List of lists" will be automatically converted to two-dimensional arrays.**

- Integers into a 2 x 2 array:

  ```python
  m = np.array([[1, 2], [3, 4]])
  m
  #> array([[1, 2],
  #>       [3, 4]])
  ```
  
<br>

--

- Floats - in this case, we explicitly specify the data type:  
  
  ```python
  m = np.array([[1, 2], [3, 4]], dtype = float)
  
  print(m)
  #> [[ 1. 2.]
  #> [ 3. 4.]]
  m.dtype.name
  #> 'float64'
  ```

---

## Initialize arrays with a given structure

- `np.zeros()` will create an array with 0s.

  We can **specify the dimensions using a tuple** &ndash;  
  here, we ask for a 3x2 array filled with floating-point zeros:

  ```python
  m = np.zeros((3, 2), dtype = float)
  m
  #> array([[ 0., 0.],
  #>        [ 0., 0.],
  #>        [ 0., 0.]])
  ```

--

- We can also "reshape" an existing array using **`reshape()`** &ndash;  
  let's create a 1-dimensional array containing the integers from 0 to 8,  
  and then re-arrange it to a 3x3 array:
  
  ```python
  a = np.arange(9)
  
  a.reshape((3, 3))
  #> array([[0, 1, 2],
  #>       [3, 4, 5],
  #>       [6, 7, 8]])
  ```

---

## Array indexing and slicing

**With one-dimensional arrays, indexing and slicing works just as we've seen
for lists, with some additional options.**

```python
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```
  
- Single-element indexing &ndash; e.g., get the 2nd element:
  
  ```python
  a[1]
  #> 1
  ```

- Create a slice &ndash; e.g., from the start to the 5th element (noninclusive):
  
  ```python
  a[:4]
  #> array([0, 1, 2, 3])
  ```
  
- Slice with negative indices &ndash; e.g., from the 3rd-to-last to the last
  element:
  
  ```python
  a[-3:]
  #> array([6, 7, 8])
  ```

---

## Array indexing and slicing

With so-called "**fancy indexing**",
which is possible with a NumPy array but not with a regular list,
we provide multiple values as a list:

```python
seqlist = list('AAGCGATTAG')
seqlist[[2, 5, 1]]
#> TypeError: list indices must be integers or slices, not list
```

```python
seqarray = np.array(list('AAGCGATTAG'))
seqarray[[2, 5, 1]]
#> array(['G', 'A', 'A'], dtype='<U1')
```

---

## Masking

Indexing with logical tests, which is sometimes called **masking**,  
can also be done on NumPy arrays thanks to vectorization.

- For example, `a > 5` will result in an array which has `True` or `False`
  for each element in the original array,
  depending on whether that number is greater than 5: 

  ```python
  a > 5
  #> array([False, False, False, False, False, False,  True,  True,  True])
  ```

- Conveniently, we can directly mask, i.e. select elements from,
  the original array with this Boolean array:

  ```python
  a[a > 5]
  #> array([6, 7, 8])
  ```

---

## Two-dimensional indexing and slicing

- Let's first create another two-dimensional array:
  
  ```python
  m = np.arange(12).reshape((3, 4))
  m
  #> array([[ 0, 1, 2, 3],
  #>        [ 4, 5, 6, 7],
  #>        [ 8, 9, 10, 11]])
  ```

- With multiple dimensions, we can separate slices for each dimension using
  commas &ndash; e.g., extract a submatrix with rows 1-2 and columns 2-4:
  
  ```python
  m[0:2, 1:4]               # m[rows, columns]
  #> array([[ 1, 2, 3],
  #>        [ 5, 6, 7]])
  ```

--

.pull-left[
- Get the **second row**:

  ```python
  m[1, :]
  #> array([4, 5, 6, 7])
  m[1]
  #> array([4, 5, 6, 7])
  ```
]

.pull-right[  
- Get the **second column**:

  ```python
  m[:, 1]
  #> array([ 1, 5, 9])
  ```
]

---

## Operate on the entire array or on rows/columns

- By default, methods like `sum()` will operate on the entire matrix:

  ```python
  m
  #> array([[ 0, 1, 2, 3],
  #>        [ 4, 5, 6, 7],
  #>        [ 8, 9, 10, 11]])
  
  m.sum()
  #> 66
  ```
  
- To get sums for each **column**, we specify **`axis = 0`**:

  ```python
  m.sum(axis = 0)
  #> array([12, 15, 18, 21])
  ```

- To get sums for each **row**, we specify **`axis = 1`**:
  
  ```python
  m.sum(axis = 1)
  #> array([ 6, 22, 38])
  ```

---

## Load data into NumPy

So far, we have mostly read data from files by creating *file handles* and then
reading data line-by-line, which is fast and saves memory.

However, with NumPy, we typically operate on multi-line arrays at once,
and therefore usually **read files into memory all at once**,
saving them as a single array object.<sup>[1]</sup>

.footnote[<sup>[1]</sup>And we'll see the same behavior for Pandas later,
with its `DataFrame` data structure.]

<br>

--

For instance, we could use NumPy's `loadtxt()` function as follows to
read a CSV file:

```python
m = np.loadtxt("my_matrix_file.csv", delimiter = ",")
```

---

## <i class="fa fa-user-edit"></i> Your turn

1. Create an array called `fives` with the elements 5, 10, 15, etc up to 45 
   using the `arange()` function:
   specify start, end, and stepsize arguments in that order.

2. Select the fourth, eight and ninth elements from `fives`
   using "fancy indexing".

2. Index/mask `fives` to only select values larger than 15.  
   Bonus: Instead, only extract values larger than 15 *and* smaller than 40.

---

## <i class="fa fa-user-edit"></i> Your turn: solutions

1. Create an array called `fives` with the elements 5, 10, 15, etc up to 45:
   
   ```python
   fives = np.arange(0, 50, 5)
   ```

2. Select the fourth, eight and ninth elements from `fives`:

   ```python
   fives[[3, 7, 8]]
   #> array([15, 35, 40])
   ```

3. Index/mask `fives` to only select values larger than 15.  
   Bonus: Instead, only extract values larger than 15 and smaller than 40.

   ```python
   fives[fives > 15]
   #> array([20, 25, 30, 35, 40, 45])
   
   fives[(fives > 15) & (fives < 40)]
   #> array([20, 25, 30, 35])
   ```

---

## <i class="fa fa-user-edit"></i> NumPy exercise

I have included the CSB section on image analysis as a "tutorial exercise" in
this week's exercises.

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Bonus material

-----

.left[
### - Random numbers and distributions (6.2.2)

### - Miscellaneous
]

<br> <br> <br> <br>

---
class: center middle inverse
name: rand

# Random numbers and distributions (6.2.2)

-----

<br> <br> <br> <br>

---
  
---
background-color: #f2f5eb

## Random numbers and distributions (6.2.2)

Being able to generate random numbers is useful to perform simulations,
randomizations, and statistics &ndash; and NumPy offers a number of functions to
get random numbers from different distributions.

As we saw last week in our popgen simulation,  
`np.random.random()` samples from a uniform distribution between 0 and 1:

```python
# import numpy as np

np.random.random(2)   # The only argument is the number of samples
#> array([ 0.31622522, 0.6173434 ])
```

---
background-color: #f2f5eb

## Random numbers and distributions (cont.)

Similarly, we can sample integers with `np.random.randint()`,  
for which we can specify not only sample size but also minimum and maximum:

```python
# If one positional argument is provided, it is the maximum,
# and sample size is 1:
np.random.randint(5)
#> 3

# If two positional arguments are provided, they are min and max,
# and sample size is 1:
np.random.randint(5, 10)
#> 8

# We can also specify the sample size:
np.random.randint(-5, -3, size = 4)
#> array([-4, -5, -5, -3])
```

--

.content-box-info[
The CSB book uses the dreadfully lengthy function name
`np.random.random_integers()` but this function has now thankfully been
deprecated in favor of `np.random.randint()`.
]

---
background-color: #f2f5eb

## Random numbers and distributions (cont.)

- `np.random.shuffle()` can be used to randomize the order of elements in an
  array:
  
  ```python
  a = np.arange(4)
  a
  #> array([0, 1, 2, 3])
  np.random.shuffle(a)     # Shuffle in place
  a
  #> array([1, 0, 2, 3])
  ```

--

<br>

- We can sample from many other distributions,
  such as the **normal distribution**,
  with parameters *mean*, *standard deviation*, and *sample size*:
  
  ```python
  np.random.normal(10, 0.1, 4)
  #> array([ 9.9574825 , 10.03459465, 9.93908575, 9.80264752])
  ```

---
background-color: #f2f5eb

## Random numbers and distributions (cont.)

Random numbers from computers are technically "pseudorandom",  
because they are generated using a deterministic algorithm.

Therefore, if we repeatedly set the **"seed"** to the same value,  
we will generate the same sequence of values (and the same as in the book!):

```python
np.random.seed(10)
print(np.random.random(1))
print(np.random.random(1))

np.random.seed(10)
print(np.random.random(1))
print(np.random.random(1))

#> [0.77132064]
#> [0.02075195]
#> [0.77132064]
#> [0.02075195]
```

--

Setting a seed can be useful to fix bugs or when you need predictable results
such as when creating a figure from the results of simulations.

---
class: center middle inverse
name: misc

# Miscellaneous

-----

<br> <br> <br> <br>

---
background-color: #f2f5eb

## Functions that work element-wise

- Square root:

  ```python
  np.sqrt(a)
  #> array([0., 1., 1.41421356, 1.73205081, 2., 2.23606798,
  #>        2.44948974, 2.64575131, 2.82842712])
```

- Exponentiation:

  ```python
  np.exp(a)
  #> array([1.00000000e+00, 2.71828183e+00, 7.38905610e+00,
  #>        2.00855369e+01, 5.45981500e+01, 1.48413159e+02,
  #>        4.03428793e+02, 1.09663316e+03, 2.98095799e+03])
  ```

---
background-color: #f2f5eb

## Initialize another array with a given structure

- As we saw last week, NumPy can generate random numbers with the `np.random`
  module &ndash; and we'll discuss several functions for this later on.

  For now, let's create an array with random values drawn from the uniform
  distribution `U[0,1]` with `np.random.random()`:

  ```python
  np.random.random((2, 3))
  #> array([[ 0.2331427 , 0.28167952, 0.66094357],
  #>       [0.13703488, 0.75519455, 0.08413554]])
  ```
  
  Note that with the tuple `(2, 3)`, we merely specify the dimensions.
