---
title: "Week 12 - Python: Regular expressions"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Regular expressions*

----

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/30

---

## Setup

```python   
import os                                       # IN PYTHON

username = os.environ.get('USER')
os.chdir('/fs/ess/PAS1855/users/' + username + '/CSB/regex/sandbox')
```

<br>

### Snakemake installation for next week

```sh
module load python/3.6-conda5.2                 # IN THE SHELL
source activate ipy-env

which pip3 # Confirm "pip3" is located inside your ipy-env:
#> ~/.conda/envs/ipy-env/bin/pip3

pip3 install snakemake
```

---

## Overview of this week

- Today: Regular expressions

- No meeting on Thursday!

---

## What are regular expressions?

> *Often, it is not feasible to search for all possible occurrences exactly as*
> *they appear in the text, but you can describe the pattern you’re looking for*
> *in your own words (e.g., find all words starting with 3 uppercase letters,*
> *followed by 4 digits).*
> ***The question is how to explain such a pattern to a computer.*
> *The answer is to use regular expressions.***
> &mdash; CSB 5.1

In other words, regular expressions are
*strings that include special symbols to succinctly describe patterns.*

---

## Regular expressions in the shell

In this course, we have already worked with regular expressions a fair bit,
using the tools `grep` and `sed` &ndash; for example:

```sh
grep -v "^#" my.gff
```

--

Exclude comment/header lines:
**`^#` matches a pound sign as the first character on a line**
(and the `-v` option inverts `grep`'s behavior: return non-matching lines).

--

```sh
echo "word1 word2" | sed -E 's/(\w+) (\w+)/\2 \1/'
```

Invert two words:

  - `\w+` matches one or more word characters (alphanumeric and `_`)
    but not the space that separates them.
  - We *capture* the two words with `()` and *refer back* to each using
    `\1` and `\2`.

--

.content-box-green[
Using regular expressions in Python is much easier because we don't have to switch
between basic/extended/Perl-like sets!
]

---

## Why use regular expressions?

- **To collect information:**  
  Match degenerate primer sequences, transcription
  factor binding sites, find accession and gene numbers, extract references
  from a manuscript.

- **As a more sophisticated replace:**  
  Replace multiple variations of a string at the same time.

- **To navigate and parse text files:**  
  Parse (semi-)structured output that
  doesn't adhere to formats for which you can use a specialized tool directly.
  
---

## The `re` module in Python

- You may recall that we used the `find()` method earlier to search within
  strings, but **`find()` only matches literal strings**.
  
- To use regular expressions, we will need to import the **`re` module**:
  
  ```python
  import re
  ```

---

## Our first match

- The main `re` function we will be using to learn about regular expression
  in Python is `re.search()`,
  which will return *the first match* it finds.<sup>[1]</sup>

  We start using it by searching for a *literal string*, after all:

  ```python
  # We define a string that we will query:
  my_string = "a given string"
  
  # We search my_string for the regex pattern "given":
  m = re.search("given", my_string)
  ```

.footnote[
<sup>[1]</sup> We'll see other functions that return all matches later.
]

---

## Match objects

- `re.search()` function returns a *match object*:

  ```python
  m
  #> <re.Match object; span=(2, 7), match='given'>
  ```
  
- On the match object, we can use the **`group()`** method to extract the
  substring that was found to match our pattern:
  
  ```python
  m.group()
  #> 'given'  # We matched a literal string so no surprises here
  ```

--

- We can also get the start and end positions (indices) of the match:

  ```python
  m.start()
  #> 2
  
  m.end()
  #> 7
  ```

---

## Failing to match

- If no matches are found, `re.search()` returns `None`:
  
  ```python
  m = re.search("9", my_string)
  print(m)
  #> None
  ```

<br>

--

- Since `None` is "**falsy**" (it will be interpreted as `False`),
  we can create a straightforward test to see if a match was made<sup> [1]</sup>:

  ```python
  if re.search("9", my_string):
        # Additional processing if a match was found
  ```

.footnote[
<sup>[1]</sup> This is a little easier than what we did last week with the
`str.find()` method,  
&nbsp; &nbsp; &nbsp; which returns `-1` if not match is found)
]


---

## Backslash plague!

Python uses backslashes `\` to give special meaning to some characters:

```python
print("my long \n line")
#> my long
#>  line
```

If we need to **include a literal backslash** in a string,
we have to escape the backslash, which we do with another backslash:

```python
print("my long \\n line")
#> my long \n line
```

--

This can get annoying, especially with regular expressions,
**which use `\` for similar purposes as Python does**:
so to match a literal backslash, we need to create a regular expression string
that contains two backslashes `\\`.  

However, we just saw that Python turns `\\` into `\`, so in fact,
we need **four** backslashes to match a single one!
*This is called the "backslash plague".*

```python
re.search('\\\\', "my long \\n line")
#> <re.Match object; span=(8, 9), match='\\'>
```

---

## "Raw string" notation

To escape this plague, Python provides raw string notation,  
which **turns off its native backslash interpretation**:

```python
print(r"my long \n line")
#> my long \n line

re.search(r"\\n", r"my long \n line")
# <re.Match object; span=(8, 10), match='\\n'>
```

So, the syntax for raw strings is to simply put an `r` before the string.

<br>

.content-box-info[
**From here on, we'll consistently use the raw string notation in regular
expressions searches.**
]

---
class: center middle inverse

# Components of regular expressions

----

<br> <br> <br>

---

## Literal characters

Literal characters can be included in regular expressions,
and as we have seen, a search pattern in `re` can consist of *only* literal characters,
but this is not very powerful.

---

## Metacharacters

| Symbol        | Negation  | Matches:
|---------------|-----------|--------
| `\n`          |           | A newline.
| `\t`          |           | A tab.    
| `\s`          | `\S`      | Any / anything but white space: space, tab, newline, carriage return.
| `\w`          | `\W`      | Any / anything but a word character: alphanumeric and underscore.
| `\d`          | `\D`      | Any / anything but a digit.
| `.`           |           | Any character.

---

## Match whitespace

- For example, to match a single whitespace character, we can use `\s`:

  ```python
  # my_string = "a given string"
  
  re.search(r"\s", my_string)
  #> <re.Match object; span=(1, 2), match=' '>
  ```

- Or we can match a whitespace character followed by five "word" characters
  (alphanumeric and underscore):

  ```python
  re.search(r"\s\w\w\w\w\w", my_string)
  #> <re.Match object; span=(1, 7), match=' given'>
  ```

.content-box-info[
Note again that `re.search()` only returns the first match.
We'll see two functions later that will return all matches.
]

---

## Character classes

You hopefully recall **character classes** from the shell,
where we used constructs like:
  - `[0-9]` to represent any single digit
  - `[ABZ]` to represent either an A, a B, or a Z (and not "*ABZ*"!).

Character classes function the same in Python
(though note that they are referred to as "*sets*" in the CSB book).

---

## Character classes (cont.)

- For example, we can search for a word that starts with a
  *lower- **or** uppercase `s`*, followed by two word (`\w`) characters:

  ```python
  my_string = "sunflowers are described on page 89"
  
  re.search(r"[sS]\w\w", my_string)
  #> <re.Match object; span=(0, 3), match='sun'>
  ```

---

## Character classes (cont.)

- Or we can search for a number consisting of two digits between 7 and 9:

  ```python
  # my_string = "sunflowers are described on page 89"
  
  re.search(r"[7-9][7-9]", my_string)
  #> <re.Match object; span=(33, 35), match='89'>
  ```

--

.content-box-q[
What would happen if `my_string` contained `898` instead of `89`?
]

--

.content-box-answer[
The match would be exactly the same.
]
  
--

.content-box-q[
What if we did not want to match such a number with three digits? 
]

--

.content-box-answer[
```python
re.search(r"[7-9][7-9]\D", my_string) # Or:
re.search(r"[7-9][7-9][^0-9]", my_string)
```
]

---

## Negating a character class

We can negate a character class by using a caret `^` as the first symbol within
the square brackets.

For example, `[^s-z]` matches anything but a letter from s-z in the alphabet: 
  
```python
# my_string = "sunflowers are described on page 89"

re.search(r"[^s-z]\w\w\w\w\w\w", my_string)
#>  <re.Match object; span=(2, 9), match='nflower'>
```

---

## Quantifiers

Instead of typing `\w` five times, like above, we can use *quantifiers* to match
multiple consecutive characters more succinctly and flexibly:

<br>

| Quantifier  | Matches |
|-------------|---------|
| **`*`**     | Preceding character *any number of times* (including 0).
| **`?`**     | Preceding character *at most* once.
| **`+`**     | Preceding character *at least* once.
| **`{n}`**   | Preceding character *exactly `n` times*.
| **`{n,}`**  | Preceding character *at least `n` times*.
| **`{n,m}`** | Preceding character *at least `n` and at most `m` times*.

---

## Using quantifiers

- From some text, we could attempt to extract a DNA sequence:
  at least three capital A/C/G/T using `[ACGT]{3,}`:
  
  ```python
  re.search(r"[ACGT]{3,}", "the motif ATTCGT").group()
  'ATTCGT'
  ```

---

## Alternations

A symbol we have also already seen is the `|` for *alternation*,
functioning as a logical or: 

```python
re.search(r"cat|mouse", "I found my cat!")
#> <re.Match object; span=(11, 14), match='cat'>

re.search(r"cat|mouse", "I found my mouse!")
#> re.search(r"cat|mouse", "I found my mouse!")
```

---

## Greedy vs. non-greedy matching

The quantifiers `?`, `*` and `+` are said to be "**greedy**".

--

This means that when faced with multiple possible endpoints of a match,
they will use the last possible endpoint.
In other words, they will **match as much text as possible**.


<br>

For example, when matching any number of characters followed by a white space
below &ndash; instead of matching `"once "`, we get:

```python
my_string = "once upon a time"
re.search(r".*\s", my_string).group()
#> 'once upon a '
```

--

Appending a question mark to the quantifier (i.e., `??`, `*?`, or `+?`)
makes the quantifier **"reluctant" or "non-greedy"** &ndash;
it will match as little as possible:

```python
re.search(r".*?\s", "once upon a time").group()
#> 'once '
```

---

## Anchors

Often, we want to restrict matching to the beginning or the end of a string  
(or equivalently, of a *line* in a file):

Anchor | Matches
-------|--------
`^`    | Beginning of the string/line (Note re-usage: **cf. `[^0-9]`**)
`$`    | End of the string/line

<br>

For example:

```python
my_string = "ATATA"

re.search(r"^TATA", my_string)
#> 

re.search(r"TATA$", my_string)
#> <re.Match object; span=(1, 5), match='TATA'>
```

---

## <i class="fa fa-user-edit"></i> Intermezzo 5.1

Describe the following regular expressions in plain English.
And what does it actually match in the `re.search()` call?

```python
re.search(r"\d" , "it takes 2 to tango").group()
```

--

Matches one digit: **`2`**.

--

<br>

```python
re.search(r"\s\w*\s", "once upon a time").group()
```

--

Matches any sequence of word characters (zero or more),
flanked by a white space on both sides: **`" upon "`**.

<br>

--

```python
re.search(r"\s\w*$", "once upon a time").group()
```

--

Matches the last word in the target string (preceded by a white space):  
**`" time"`**.

---

## What if I need to match a literal `*`?

Sometimes, you need to **literally match a special character**,
like a `*`.  
That is, you want to take away ("escape") its special meaning.

How can we do this?

--

Just like escaping backslashes,
we can escape any other character's special meaning **with a backslash**:

```python
re.search(r'*', 'my*string')
#> error: nothing to repeat at position 0

re.search(r'\*', 'my*string')
#> <re.Match object; span=(2, 3), match='*'>
```

.content-box-info[
We also saw this in the shell where one way to match a file name with a space,
other than quoting, is to escape the space:

```sh
$ ls my\ bad\ file
#> 'my bad file'
```
]

---

## Other functions of the *re* module

Function     | Behavior 
-------------|---------
`re.findall(reg,target)`  | As `re.search`, but return a **list** of all the matches.
`re.finditer(reg,target)` | As `re.findall`, but return an **iterator**.
`re.compile(reg)`          | Compile a regular expression. <br> This stores the pattern for repeated use, improving the speed.
`re.split(reg,target)`    | Split the text by the occurrence of the pattern described by the regular expression.
`re.sub(reg,repl,target)` | Substitute each non-overlapping occurrence of the match with the text in `repl`.

---

## Example: Querying a FASTA file

Say that we want to find methylation sites in nucleotide sequences from
*Escherichia coli* that are stored in the file `CSB/regex/data/Ecoli.fasta`.

There are several methylases in *E. coli*.
The *EcoKI* methylase, for example, modifies the sequences `AACNNNNNNGTGC` *and*
`GCACNNNNNNGTT`.

--

Let's take a look at the file:

```python
fafile = "../data/Ecoli.fasta"

!head -5 {fafile}
#> >gi|556503834|ref|NC_000913.3|:1978338-2028069 Escherichia coli str. K-12 substr. #> MG1655, complete genome
#> AATATGTCCTTACAAATAGAAATGGGTCTTTACACTTATCTAAGATTTTTCCTAAATCGACGCAACTGTA
#> CTCGTCACTACACGCACATACAACGGAGGGGGGCTGCGATTTTCAATAATGCGTGATGCAGATCACACAA
#> AACACTCAATTACTTAACATAAATGGGTAATGACTCCAACTTATTGATAGTGTTTTATGTTCAGATAATG
#> CCCGATGACTTTGTCATGCAGCTCCACCGATTTTGAGAACGACAGCGACTTCCGTCCCAGCCGTGCCAGG

!grep ">" {fafile}
#> >gi|556503834|ref|NC_000913.3|:1978338-2028069 Escherichia coli str. K-12 substr. MG1655, complete genome
#> >gi|556503834|ref|NC_000913.3|:4035299-4037302 Escherichia coli str. K-12 substr. MG1655, complete genome
```

---

## Example: Querying a FASTA file (cont.)

Let's read the FASTA file, and take the sequence for the first of the two
records in the file:

```python
from Bio import SeqIO
genes = list(SeqIO.parse(fafile, "fasta"))
seq1 = str(genes[0].seq)
```

```python
len(seq1)
#> 49732
seq1[:40]
#> 'AATATGTCCTTACAAATAGAAATGGGTCTTTACACTTATC'
```

<br> <br>

.content-box-info[
Note, we used with `SeqIO` instead of `pyfaidx` (as in CSB) here to simplify
the sequence parsing and not introduce a new module.
]

---

## Example: Querying a FASTA file (cont.)

Since our regular expression is quite complicated,
we will first save it using the `re.compile()` function:

```python
EcoKI = re.compile(r"AAC[ATCG]{6}GTGC|GCAC[ATCG]{6}GTT")

# Matches:
# AACNNNNNNGTGC
# GCACNNNNNNGTT
```

---

## Example: Querying a FASTA file (cont.)

Also, we will need a different `re` function to get *all* matches.  
Let's first try **`re.findall()`**:

```python
re.findall(EcoKI, seq1)
#> ['AACAGCATCGTGC', 'AACTGGCGGGTGC', 'GCACCACCGCGTT',
#>  'GCACAACAAGGTT', 'GCACCGCTGGGTT', 'AACCTGCCGGTGC']
```

That simply returned a list of matches, but we want the *positions* too!

--

Instead, we can use **`re.finditer()`**,
which will get us match objects like we got for `re.search()` earlier,
that we can iterate over:

```python
for match in re.finditer(EcoKI, seq1):
    print(match.start() + 1, match.group())
#> 18452 AACAGCATCGTGC
#> 18750 AACTGGCGGGTGC
#> 25767 GCACCACCGCGTT
#> 35183 GCACAACAAGGTT
#> 40745 GCACCGCTGGGTT
#> 42032 AACCTGCCGGTGC
```

---

## Groups and backreferences in regular expressions

Using parentheses, we can create **groups** in regular expressions.

We have already seen the use of such groups with backreferences in `sed`,  
and another feature of groups is that **quantifiers will operate on the entire
group**:

```python
re.search(r"GT{2}", "ATGGTGTCCGTGTT").group()
#> 'GTT'
re.search(r"(GT){2}", "ATGGTGTCCGTGTT").group()
#> 'GTGT'
```

---

## Example: retrieving the sequence between primers

Say that we want to identify the variable region that
is flanked by the bacterial ribosomal RNA primers 799F and 904R.

We can build a regular expression with the two primers,
each of which we capture with `()`, and the unknown sequence which we also
capture:

```python
primer1 = 'AAC[AC]GGATTAGATACCC[GT]G'
primer2 = '[CT]T[AG]AAACTCAAATGAATTGACGGGG'
regpat = '(' + primer1 + ')' + '([ATCG]+)' + '(' + primer2 + ')'

regpat
#> '(AAC[AC]GGATTAGATACCC[GT]G)([ATCG]+)([CT]T[AG]AAACTCAAATGAATTGACGGGG)'
```

--

We will search for this pattern in the second sequence of `Ecoli.fasta`:

```python
# from Bio import SeqIO 
# genes = list(SeqIO.parse("../data/Ecoli.fasta", "fasta"))
seq2 = str(genes[1].seq)

m = re.search(regpat, seq2)
```

---

## Example: retrieving the sequence between primers

Thanks to our groupings, we can now refer back to these separately:

- The entire match with `group(0)`:

  ```python
  len(m.group(0))
  #> 148
  ```

- Our forward and reverse primers with `group(1)` and `group(3)`

  ```python
  m.group(1)
  #> 'AACAGGATTAGATACCCTG'
  m.group(3)
  #> 'TTAAAACTCAAATGAATTGACGGGG'
  ```

- And the sequence of interest with `group(2)`:

  ```python
  m.group(2)
  #> 'GTAGTCCACGCCGTAAACGATGTCGACTTGGAGGTTGTGCCCTTGAGGCGTGGCTTCCGGAGCTAACGCGTTAAGTCGACCGCCTGGGGAGTACGGCCGCAAGG'
  len(m.group(2))
  #> 104
  ```

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---

## Example: finding names of MHC alleles in BLAST results

The file `Marra2014_BLAST_data.txt` has some BLAST results in tabular format.
We are looking for MHC alleles and we happen to know that their names are 
unique in containing an asterisk `*`.

We start by moving to the right directory and peeking at the file:

```python
import os
os.chdir('/fs/ess/PAS1855/users/<user>/CSB/regex/sandbox')

!head ../data/Marra2014_BLAST_data.txt
#> Seq. Name	Seq. Description	Seq. Length	#Hits	min. eValue	mean Similarity
#> contig00001	---NA---	112	0		
#> contig00002	pol2_mouse ame: full=retrovirus-related pol polyprotein line-1 ame: full=long interspersed element-1 short=l1 includes: ame: full=reverse transcriptase includes: ame: full=endonuclease	932	3	6.53E-113	79.67%
#> contig00003	lin1_nycco ame: full=line-1 reverse transcriptase homolog	2070	5	0	59.60%
#> contig00004	lin1_nycco ame: full=line-1 reverse transcriptase homolog	3074	5	6.92E-68	64.80%
#> contig00005	---NA---	190	0
```

---

## Example: finding names of MHC alleles in BLAST results

Let's first find, print and count lines with an `*`:

```python
with open("../data/Marra2014_BLAST_data.txt") as f:
    counter = 0
    for line in f:
        m = re.search(r"\*", line) # Search for '*' in each line
        if m:             # If a match was found:
            counter += 1  # Increment counter
            print(line)   # Print the line
            
#> contig01987	2b14_human ame: full=hla class ii histocompatibility drb1-4 beta chain ame: full=mhc class ii antigen drb1*4 short=dr-4 short=dr4 flags: precursor	112	5	5.07E-10	78.60%
#> contig05816	1c08_human ame: full=hla class i histocompatibility cw-8 alpha chain ame: full=mhc class i antigen cw*8 flags: precursor	825	5	1.06E-97	67.80%
#> contig05821	1b46_human ame: full=hla class i histocompatibility b-46 alpha chain ame: full=bw-46 ame: full=mhc class i antigen b*46 flags: precursor	606	5	9.86E-09	67.60%
#> contig22120	2b11_human ame: full=hla class ii histocompatibility drb1-1 beta chain ame: full=mhc class ii antigen drb1*1 short=dr-1 short=dr1 flags: precursor	137	5	5.05E-11	76.80%
#> contig23154	mmrn1_human ame: full=multimerin-1 ame: full=emilin-4 ame: full=elastin microfibril interface located protein 4 short=elastin microfibril interfacer 4 ame: full=endothelial cell multimerin contains: ame: full=platelet glycoprotein ia* contains: ame: full=155 kda platelet multimerin short=p-155 short=p155 flags: precursor	4524	5	0	62.00%
#> contig27667	ud13_human ame: full=udp-glucuronosyltransferase 1-3 short=udpgt 1-3 short=ugt1*3 short=ugt1-03 short= ame: full=udp-glucuronosyltransferase 1-c short=ugt-1c short=ugt1c ame: full=udp-glucuronosyltransferase 1a3 flags: precursor	1818	5	0	84.40%
```

<br>

```python
print("The pattern was matched in", counter, "lines")
#> The pattern was matched in 6 lines
```

---

## Example: finding names of MHC alleles in BLAST results

```sh
#> [...] full=mhc class ii antigen drb1*1 short=dr-1 [...]
#> [...] full=mhc class i antigen cw*8 flags: precursor 606 [...]
```

From looking at those lines, if we want to extract just the names of the MHC
alleles, we should start matching at `mhc` and stop matching after the *word*
that contains the asterisk `*`.

So, our regex could be: `mhc[\s\w*]+\*\w*`:

Component    | Meaning 
-------------|---------
`mhc`        | Match the literal characters `mhc`.
`[\s\w*]+`   | Match a white space or zero or more word characters, one or more times.
`\*`         | Match a (literal) asterisk.
`\w*`        | Match zero or more word characters.



---

## Example: finding names of MHC alleles in BLAST results

Let's try it:

```python
with open("../data/Marra2014_BLAST_data.txt") as f:
    for line in f:
        m = re.search(r"mhc[\s\w*]+\*\w*", line)
        if m:
            print(m.group())

#> mhc class ii antigen drb1*4
#> mhc class i antigen cw*8
#> mhc class i antigen b*46
#> mhc class ii antigen drb1*1
```

---

## <i class="fa fa-user-edit"></i> More intermezzo 5.1

```python
re.search(r"\s\w{1,3}\s", "once upon a time").group()
```

--

Matches a sequence of one to three word characters,
flanked by two white spaces: `" a "`.

```python
re.search(r"\w*\s\d.*\d", "take 2 grams of H2O").group()
```

--

Matches zero or more word characters (`\w*`),
followed by a white space (`\s`),
followed by a digit (`\d`),
zero or more characters (`.*`),
and ending with a digit (`\d`): `take 2 grams of H2`.

---

## <i class="fa fa-user-edit"></i> Intermezzo 5.2

Let’s practice translating from plain English to regular expressions.

The NCBI GenBank contains information on nucleotide sequences,
protein sequences, and whole genome sequences (WGS). 

The following table describes the construction of sequence identifiers in plain English. Construct the appropriate regular expression to match either protein, WGS or nucleotide IDs:

1. **Protein**: 3 letters + 5 numerals

2. **WGS**: 4 letters + 2 numerals for WGS assembly version + 6–8 numerals

3. **Nucleotide**: 1 letter + 5 numerals OR 2 letters + 6 numerals

---

## <i class="fa fa-user-edit"></i> Intermezzo 5.2: Solutions

1. 
```python
r"[A-Za-z]{3}\d{5}"
```

<br>

2. 
```python
r"[A-Za-z]{4}\d{8,10}"
```

<br>

3. 
```python
r"([A-Z]{1}\d{5}|[A-Z]{2}\d{6})"
```
