---
title: "Week 11 - Python: Scientific computing"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
#xaringanExtra::use_scribble()
#xaringanExtra::use_panelset()
#xaringanExtra::use_clipboard()
#xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Scientific computing*

----

# I (Tuesday)

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/23 (updated: `r Sys.Date()`)

---

## Overview of this week

.pull-left[
**This presentation**

- NumPy and SciPy (CSB 6.2)

]

.pull-right[
**Thursday's presentation:**

- Pandas (CSB 6.3)

- BioPython (CSB 6.4)
]

<br> <br>

---
class: center middle inverse

# NumPy and SciPy (CSB 6.2)

-----

<br> <br> <br> <br>

---

## What are NumPy and Scipy?

NumPy is the basic Python package for scientific computing,
which includes efficient data structures for vectors and matrices.

SciPy is a collection of statistical and mathematical functions meant to
extend the features of NumPy.

<br> <br>

--

We have already used SciPy last week, to generate random numbers,
and will start with NumPy now.
NumPy is typically imported `as np`, so we can access its functions with a
little less typing:

```python
import numpy as np
```

---

## NumPy arrays

Arrays are NumPy's main data structure,
which can contain numerical vectors (1-dimensional arrays),
matrices (2-dimensional arrays),
and tensors (n-dimensional arrays).

To create a one-dimensional array,
we can use NumPy's variant on the `range()` function, `np.arange()`:

```python
a = np.arange(9)
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```

---

## NumPy arrays (cont.)

- Recall that with lists, we can use operators like `*` but they are not
  applied element-wise:

  ```python
  mylist = list(range(9))
  mylist
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8]
  
  mylist * 2
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8, 0, 1, 2, 3, 4, 5, 6, 7, 8]
  ```

--

- However, with arrays we can easily do element-wise ("vectorized") operations:

  ```python
  # Multiply each element by 2:
  a * 2
  #> array([ 0, 2, 4, 6, 8, 10, 12, 14, 16])
  
  # Add 2 to each element:
  2 + a
  #> array([ 2, 3, 4, 5, 6, 7, 8, 9, 10])
  ```

---

## Some functions and methods for arrays

- Get the length (number of elements) along each dimension with the `shape()`
  method:

  ```python
  a.shape
  (9,)
  ```

- Get the number of dimensions with the `ndim()` method:

  ```python
  a.ndim
  #? 1
  ```

- Get the number of elements with the `size()` method:
  
  ```python
  a.size
  #> 9
  ```

- Get the data type for the elements in the array
  (64-bit integers):
  
  ```python
  a.dtype.name
  #> 'int64'
  ```

---

## Simple Arithmetic and statistics for arrays

```python
# Sum of all elements:
a.sum()
#> 36

# Mean:
a.mean()
#> 4.0

# Standard deviation:
a.std()
#> 2.5819888974716112

# Lowest value:
a.min()
#> 0

# Highest value:
a.max()
#> 8
```

---

## Functions that work element-wise

- Square root:

  ```python
  np.sqrt(a)
  #> array([0., 1., 1.41421356, 1.73205081, 2., 2.23606798,
  #>        2.44948974, 2.64575131, 2.82842712])
```

- Exponentiation:

  ```python
  np.exp(a)
  #> array([1.00000000e+00, 2.71828183e+00, 7.38905610e+00,
  #>        2.00855369e+01, 5.45981500e+01, 1.48413159e+02,
  #>        4.03428793e+02, 1.09663316e+03, 2.98095799e+03])
  ```

---

## Creating arrays by converting lists

Convert a list to a one-dimensional array:

- Integers:

  ```python
  a1 = np.array([1, 2, 3, 4])
  print(a1)
  #> [1 2 3 4]
  
  a1.dtype.name
  #> 'int64'
  ```

- Floats:

  ```python
  a1 = np.array([1.0, 2.0, 3.0, 4.0])
  
  print(a1)
  #> [ 1. 2. 3. 4.]
  a1.dtype.name
  #> 'float64'
  ```

---

## Creating arrays by converting lists (cont.)

Convert a list of lists to a two-dimensional array.

- Integers:

  ```python
  m = np.array([[1, 2], [3, 4]])
  m
  #> array([[1, 2],
  #>       [3, 4]])
  ```
  
- Floats - now we explicitly specify the data type:  
  
  ```python
  m = np.array([[1, 2], [3, 4]], dtype = float)
  
  print(m)
  #> [[ 1. 2.]
  #> [ 3. 4.]]
  m.dtype.name
  #> 'float64'
  ```

---

## Initialize arrays with a given structure

- create 3x2 array filled with floating-point zeros:

  ```python
  m = np.zeros((3, 2), dtype = float)
  m
  #> array([[ 0., 0.],
  #> [ 0., 0.],
  #> [ 0., 0.]])
  ```

- Create an array containing the integers from 0 to 8,
  and arrange it in a 3x3 array:
  
  ```python
  a = np.arange(9)
  a.reshape((3,3))
  #> array([[0, 1, 2],
  #>       [3, 4, 5],
  #>       [6, 7, 8]])
  ```

---

## Initialize arrays with a given structure (cont.)

- Create an array with random values,
  drawn from uniform distribution U[0,1]

  ```python
  np.random.random((2, 3))
  #> array([[ 0.2331427 , 0.28167952, 0.66094357],
  #>       [0.13703488, 0.75519455, 0.08413554]])
```

---

## Array indexing

With one-dimensional arrays, indexing works just like we saw for lists:

```python
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```
  
- Single-element indexing - 2nd element:
  
  ```python
  a[1]
  #> 1
  ```

- Create a slice - start to the 5th element (noninclusive):
  
  ```python
  a[:4]
  #> array([0, 1, 2, 3])
  ```
  
- Slice with negative indices - 3rd to last to last element:
  
  ```python
  a[-3:]
  #> array([6, 7, 8])
  ```
---

## Two-dimensional indexing

```python
m = np.arange(16).reshape((4, 4))
m
#> array([[ 0, 1, 2, 3],
#>        [ 4, 5, 6, 7],
#>        [ 8, 9, 10, 11],
#>        [12, 13, 14, 15]])
```

- Extract submatrix:
  
  ```python
  m[0:3, 1:4]
  #> array([[ 1, 2, 3],
  #>        [ 5, 6, 7],
  #>        [ 9, 10, 11]])
  ```
  
- Get the second row:

  ```python
  m[1]
  #> array([4, 5, 6, 7])
  ```
  
- Get the second column:

  ```python
  m[:, 1]
  #> array([ 1, 5, 9, 13])
  ```

---

## Operate on the whole array, rows, or columns

- Whole matrix operation:

  ```python
  m.sum()
  #> 120
  ```
  
- Column sums:

  ```python
  m.sum(axis = 0)
  #> array([24, 28, 32, 36])
  ```

- Row sums:
  
  ```python
  m.sum(axis = 1)
  #> array([ 6, 22, 38, 54])
  ```

---

## Load data into NumPy

```python
m =
np.loadtxt("my_matrix_file.csv", delimiter = ",")
```

---

## Example: image processing with NumPy


---

## Random numbers and distributions (6.2.2)

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>


