---
title: "Week 11 - Python: Scientific computing"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
#xaringanExtra::use_scribble()
#xaringanExtra::use_panelset()
#xaringanExtra::use_clipboard()
#xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Scientific computing*

----

# I: NumPy

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/23 (updated: `r Sys.Date()`)

---

## Overview of this week

.pull-left[
**This presentation**

- NumPy and SciPy (CSB 6.2)

]

.pull-right[
**Thursday's presentation:**

- Pandas (CSB 6.3)

- BioPython (CSB 6.4)
]

<br> <br>

---
class: center middle inverse

# NumPy and SciPy (CSB 6.2)

-----

<br> <br> <br> <br>

---

## Setup

```sh
sinteractive -A PAS1855 -t 60
module load python/3.6-conda5.2
source activate ipy-env
conda install -y scikit-image
```

---

## What are NumPy and Scipy?

NumPy is the basic Python package for scientific computing,
which includes efficient data structures for vectors and matrices.

SciPy is a collection of statistical and mathematical functions meant to
extend the features of NumPy.

<br> <br>

--

We have already used NumPy a bit last week to generate random numbers.
As we saw, NumPy is typically imported with `as np` added,
so we can access its functions with a little less typing:

```python
import numpy as np
```

---

## NumPy arrays

Arrays are NumPy's main data structure,
which can contain numerical **vectors** (1-dimensional arrays),
**matrices** (2-dimensional arrays), and **tensors** (n-dimensional arrays).

To create a one-dimensional array,
we can use NumPy's variant on the `range()` function, `np.arange()`:

```python
a = np.arange(9)
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```

---

## NumPy arrays (cont.)

- Recall once again that with lists,
  we can use operators like `*` but they are not applied element-wise:

  ```python
  mylist = list(range(9))
  mylist
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8]
  
  mylist * 2
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8, 0, 1, 2, 3, 4, 5, 6, 7, 8]
  ```

--

- However, with arrays, these operators do work **element-wise**
  (are "vectorized"):

  ```python
  # Multiply each element by 2:
  a * 2
  #> array([ 0, 2, 4, 6, 8, 10, 12, 14, 16])
  
  # Add 2 to each element:
  2 + a
  #> array([ 2, 3, 4, 5, 6, 7, 8, 9, 10])
  ```

---

## Some attributes of arrays

- Get the length (number of elements) along each dimension with `shape`:

  ```python
  a.shape           # NOTE: No parentheses!
  #> (9,)           # We have only one dimension in this case
  ```

- Get the number of dimensions with `ndim`:

  ```python
  a.ndim
  #> 1
  ```

--

- Get the number of elements with the `size`:
  
  ```python
  a.size
  #> 9
  ```

- Get the data type for the elements in the array:
  
  ```python
  a.dtype.name
  #> 'int64'        # 64-bit integers
  ```

---

## Simple Arithmetic and statistics for arrays

- Sum of all elements with `sum()`:

  ```python
  a.sum()
  #> 36
  ```
  
- Mean and standard deviation with `mean()` and `std()`:
  
  ```python
  a.mean()
  #> 4.0

  a.std()
  #> 2.5819888974716112
  ```

--

- Lowest and highest value with `min()` and `max()`:

  ```python
  a.min()
  #> 0

  a.max()
  #> 8
  ```

---

## Functions that work element-wise

- Square root:

  ```python
  np.sqrt(a)
  #> array([0., 1., 1.41421356, 1.73205081, 2., 2.23606798,
  #>        2.44948974, 2.64575131, 2.82842712])
```

- Exponentiation:

  ```python
  np.exp(a)
  #> array([1.00000000e+00, 2.71828183e+00, 7.38905610e+00,
  #>        2.00855369e+01, 5.45981500e+01, 1.48413159e+02,
  #>        4.03428793e+02, 1.09663316e+03, 2.98095799e+03])
  ```

---

## Creating arrays by converting lists

**Converting a list to a one-dimensional array**

- If we create an array of numbers without decimal points,  
  they will be interpreted as integers:

  ```python
  a1 = np.array([1, 2, 3, 4])
  print(a1)
  #> [1 2 3 4]
  
  a1.dtype.name
  #> 'int64'
  ```

- If we include decimal points, they will be floats:

  ```python
  a1 = np.array([1.0, 2.0, 3.0, 4.0])
  
  print(a1)
  #> [ 1. 2. 3. 4.]
  a1.dtype.name
  #> 'float64'
  ```

---

## Creating arrays by converting lists (cont.)

**Converting a "list of lists" to a two-dimensional array.**

- Integers:

  ```python
  m = np.array([[1, 2], [3, 4]])
  m
  #> array([[1, 2],
  #>       [3, 4]])
  ```
  
- Floats - now we explicitly specify the data type:  
  
  ```python
  m = np.array([[1, 2], [3, 4]], dtype = float)
  
  print(m)
  #> [[ 1. 2.]
  #> [ 3. 4.]]
  m.dtype.name
  #> 'float64'
  ```

---

## Initialize arrays with a given structure

- `np.zeros()` will create an array with 0s.

  We can **specify the dimensions using a tuple** &ndash;  
  here, we ask for a 3x2 array filled with floating-point zeros:

  ```python
  m = np.zeros((3, 2), dtype = float)
  m
  #> array([[ 0., 0.],
  #> [ 0., 0.],
  #> [ 0., 0.]])
  ```

--

- We can also "reshape" an existing array using `reshape()` &ndash;  
  let's create a 1-dimensional array containing the integers from 0 to 8,  
  and then re-arrange it to a 3x3 array:
  
  ```python
  a = np.arange(9)
  
  a.reshape((3,3))
  #> array([[0, 1, 2],
  #>       [3, 4, 5],
  #>       [6, 7, 8]])
  ```

---

## Array indexing

**With one-dimensional arrays, indexing works just as we've seen for lists.**

```python
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```
  
- Single-element indexing &ndash; e.g., get the 2nd element:
  
  ```python
  a[1]
  #> 1
  ```

- Create a slice &ndash; e.g., from the start to the 5th element (noninclusive):
  
  ```python
  a[:4]
  #> array([0, 1, 2, 3])
  ```
  
- Slice with negative indices &ndash; e.g., from the 3rd-to-last to the last
  element:
  
  ```python
  a[-3:]
  #> array([6, 7, 8])
  ```
---

## Two-dimensional indexing

```python
m = np.arange(16).reshape((4, 4))
m
#> array([[ 0, 1, 2, 3],
#>        [ 4, 5, 6, 7],
#>        [ 8, 9, 10, 11],
#>        [12, 13, 14, 15]])
```

- With multiple dimensions, we can separate slices for each dimension using
  commas &ndash; extract a submatrix with rows 1-3 and columns 2-4:
  
  ```python
  m[0:3, 1:4]               # m[rows, columns]
  #> array([[ 1, 2, 3],
  #>        [ 5, 6, 7],
  #>        [ 9, 10, 11]])
  ```

--

.pull-left[
- Get the second row:

  ```python
  m[1, :]
  #> array([4, 5, 6, 7])
  m[1]
  #> array([4, 5, 6, 7])
  ```
]

.pull-right[  
- Get the second column:

  ```python
  m[:, 1]
  #> array([ 1, 5, 9, 13])
  ```
]

---

## Operate on the whole array, rows, or columns

- By default, methods like `sum()` will operate on the entire matrix:

  ```python
  m.sum()
  #> 120
  ```
  
<br>

- To get sums for each **column**, we specify **`axis = 0`**:

  ```python
  m.sum(axis = 0)
  #> array([24, 28, 32, 36])
  ```

- To get sums for each **row**, we specify **`axis = 1`**:
  
  ```python
  m.sum(axis = 1)
  #> array([ 6, 22, 38, 54])
  ```

---

## Load data into NumPy

Rather than creating *file handles* and reading data line-by-line,
like we did in the past few weeks, NumPy (as well as Pandas) typically
**read files into memory all at once, and save the contents in an object**.

<br>

For instance, we can use NumPy's `loadtxt()` function as follows:

```python
m = np.loadtxt("my_matrix_file.csv", delimiter = ",")
```

---

## Case study: image processing with NumPy

- We start by loading NumPy and Scikit-image:

  ```python
  #import numpy as np        # Should be loaded already
  import skimage.io as io
  ```

- Next, let's save the path to the image we want to read in as a variable:
  ```python
  import os
  
  CSB_dir = '/fs/ess/PAS1855/users/<user>/CSB/scientific/data'
  # CSB_dir = '../CSB/scientific/data'
  
  image_file = 'Kacsoh2013_Drosobrain.png'
  image_path = os.path.join(CSB_dir, image_file)
  ```

--

.content-box-info[
If using a Jupyter Notebook, we would also need to make Matplotlib image plotting
available using this IPython "magic function":

```python
%matplotlib inline
```
]

---

## Case study: Image processing with NumPy (cont.)

Let's view the image:

```python
image = io.imread(image_path)
io.imshow(image)
```

<p align="center">
<img src=img/red.png width="55%">
</p>

---

## Case study: Image processing with NumPy (cont.)

- How is our image saved?

  ```python
  type(image)
  #> numpy.ndarray
  ```
  
  As a NumPy array!

<br>

--

- Therefore, we can use the NumPy methods we have learned about.
  Let's check the dimensions:
  
  ```python
  image.shape
  (1024, 1024, 4)
  ```
  
  The image is 1024 x 1024 pixels, with RGBA along the 3rd dimension:
    - Red (index 0)
    - Green (index 1)
    - Blue (index 2)
    - Alpha (opaqueness; index 3).

---

## Case study: Image processing with NumPy (cont.)

- The first channel is the red channel, which should be the one of interest,
  given what our image looks like... Let's extract it and explore a bit:
  
  ```python
  red = image[:, :, 0]
  
  print(red.mean())
  #> 23.181024551391602
  print(red.std())
  #> 30.446740821993011
  print(red.min(), red.max())
  #> 0 255
  ```

--

- It looks like we indeed *only* have variation in the red channel:

  ```python
  for channel_index in range(0, image.shape[2]):
        print(image[:, :, channel_index].std())
  #> 30.44674082199301
  #> 0.0
  #> 0.0
  #> 0.0
  ```

---

## Case study: Image processing with NumPy (cont.)

Say that we are interested in examining what area in the image express the
NPF and NPFR1 genes that were targeted by this assay.

We'll here take an overly simple approach,
by **setting a threshold of 100** and determining which pixels have a higher
value than that.

---

## Case study: Image processing with NumPy (cont.)

Let's check what a value of 100 looks like by modifying a copy of the image:

  ```python
  img_copy = image.copy()
  img_copy[:, 480:500, 0] = 100
  io.imshow(img_copy)
  ```
  
<p align="center">
<img src=img/redline.png width="50%">
</p>

---

## Case study: Image processing with NumPy (cont.)

- Next, we'll create a "mask", an array of that indicates whether the pixel
  intensity of the red channel is above the threshold.

  ```python
  threshold = 100
  # red > threshold produces a Boolean array - times 1 to get 0/1
  mask = (red > threshold) * 1
  ```

--

.pull-left[

How many pixels pass?

```python
mask.sum()
#> 37037
```

Plot the mask:

```python
io.imshow(mask)
```

]

.pull-right[
<p align="left">
<img src=img/mask1.png width="100%">
</p>
]

---

## Case study: Image processing with NumPy (cont.)

Finally, we'll use a little trick to get the actual pixel values back for those
pixels that passed the threshold &ndash; multiply by the mask:

```python
mask2 = red * mask
io.imshow(mask2)
```

<p align="center">
<img src=img/mask2.png width="60%">
</p>

---

## Random numbers and distributions (6.2.2)

Being able to generate random numbers is useful to perform simulations,
randomizations, and statistics &ndash; and NumPy offers a number of functions to
get random numbers from different distributions.

As we saw earlier `np.random.random()` will by sample from a uniform
distribution between 0 and 1:

```python
# import numpy as np

np.random.random(2)   # The only argument is the number of samples
#> array([ 0.31622522, 0.6173434 ])
```

---

## Random numbers and distributions (cont.)

Similarly, we can sample integers with `np.random.randint()`,  
for which we can specify not only sample size but also minimum and maximum:

```python
# If one positional argument is provided, it is the maximum:
np.random.randint(5)
#> 3

# If two positional arguments are provided, they are min and max:
np.random.randint(5, 10)
#> 8

# We can also specify the sample size:
np.random.randint(-5, -3, size = 4)
#> array([-4, -5, -5, -3])
```

--

.content-box-info[
The CSB book uses the truly dreadfully named `np.random.random_integers()`
but this function has now thankfully been deprecated in favor of
`np.random.randint()`.
]

---

## Random numbers and distributions (cont.)

- `np.random.shuffle()` can be used to randomize the order of elements in an
  array:
  
  ```python
  a = np.arange(4)
  a
  #> array([0, 1, 2, 3])
  np.random.shuffle(a)     # Shuffle in place
  a
  #> array([1, 0, 2, 3])
  ```

---

## Random numbers and distributions (cont.)

We can sample from many other distributions, for instance:

- **Beta**, with parameters alpha, beta, and sample size:

  ```python
  np.random.beta(1/2, 1/2, 4)
  array([ 0.64371741, 0.07697382, 0.04568635, 0.04068809])
  ```

- **Normal**, with parameters: mean, standard deviation, and sample size:
  
  ```python
  np.random.normal(10, 0.1, 4)
  #> array([ 9.9574825 , 10.03459465, 9.93908575, 9.80264752])
  ```

---

## Random numbers and distributions (cont.)

Random numbers from computers are technically "pseudorandom" because they are
generated using a deterministic algorithm.

Therefore, if we set the **"seed"** to the same value,
we will generate the same values (and the same as in the book too!):

```python
np.random.seed(10)
print(np.random.random(1))
print(np.random.random(1))

np.random.seed(10)
print(np.random.random(1))
print(np.random.random(1))

#> [0.77132064]
#> [0.02075195]
#> [0.77132064]
#> [0.02075195]
```

--

Setting a seed can be useful to fix bugs or when you need predictable results
such as when creating a figure from the results of simulations.

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---

## Initialize arrays with a given structure (cont.)

- As we saw last week, NumPy can generate random numbers with the `np.random`
  module &ndash; and we'll discuss several functions for this later on.

  For now, let's create an array with random values drawn from the uniform
  distribution `U[0,1]` with `np.random.random()`:

  ```python
  np.random.random((2, 3))
  #> array([[ 0.2331427 , 0.28167952, 0.66094357],
  #>       [0.13703488, 0.75519455, 0.08413554]])
  ```
  
  Note that with the tuple `(2, 3)`, we merely specify the dimensions.

---

.content-box-info[
### Magic commands

]
