---
title: "Week 11 - Python: Scientific computing"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
#xaringanExtra::use_scribble()
#xaringanExtra::use_panelset()
#xaringanExtra::use_clipboard()
#xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Scientific computing*

----

# I (Tuesday)

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/23 (updated: `r Sys.Date()`)

---

## Overview of this week

.pull-left[
**This presentation**

- NumPy and SciPy (CSB 6.2)

]

.pull-right[
**Thursday's presentation:**

- Pandas (CSB 6.3)

- BioPython (CSB 6.4)
]

<br> <br>

---
class: center middle inverse

# NumPy and SciPy (CSB 6.2)

-----

<br> <br> <br> <br>

---

## What are NumPy and Scipy?

NumPy is the basic Python package for scientific computing,
which includes efficient data structures for vectors and matrices.

SciPy is a collection of statistical and mathematical functions meant to
extend the features of NumPy.

<br> <br>

--

We have already used SciPy last week, to generate random numbers,
and will start with NumPy now.
NumPy is typically imported with `as np` added,
so we can access its functions with a little less typing:

```python
import numpy as np
```

---

## NumPy arrays

Arrays are NumPy's main data structure,
which can contain numerical vectors (1-dimensional arrays),
matrices (2-dimensional arrays),
and tensors (n-dimensional arrays).

To create a one-dimensional array,
we can use NumPy's variant on the `range()` function, `np.arange()`:

```python
a = np.arange(9)
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```

---

## NumPy arrays (cont.)

- Recall that with lists, we can use operators like `*` but they are not
  applied element-wise:

  ```python
  mylist = list(range(9))
  mylist
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8]
  
  mylist * 2
  #>[0, 1, 2, 3, 4, 5, 6, 7, 8, 0, 1, 2, 3, 4, 5, 6, 7, 8]
  ```

--

- However, with arrays, these operarors do perform **element-wise** ("vectorized")
  operations:

  ```python
  # Multiply each element by 2:
  a * 2
  #> array([ 0, 2, 4, 6, 8, 10, 12, 14, 16])
  
  # Add 2 to each element:
  2 + a
  #> array([ 2, 3, 4, 5, 6, 7, 8, 9, 10])
  ```

---

## Some attributes of arrays

- Get the length (number of elements) along each dimension with `shape`:

  ```python
  a.shape           # NOTE: No parentheses!
  #> (9,)           # We have only one dimension in this case
  ```

- Get the number of dimensions with `ndim`:

  ```python
  a.ndim
  #> 1
  ```

- Get the number of elements with the `size`:
  
  ```python
  a.size
  #> 9
  ```

- Get the data type for the elements in the array:
  
  ```python
  a.dtype.name
  #> 'int64'        # 64-bit integers
  ```

---

## Simple Arithmetic and statistics for arrays


- Sum of all elements with `sum()`:

  ```python
  a.sum()
  #> 36
  ```
  
- Mean and standard deviation with `mean()` and `std()`:
  
  ```python
  a.mean()
  #> 4.0

  a.std()
  #> 2.5819888974716112
  ```

- Lowest and highest value with `min()` and `max()`:

  ```python
  a.min()
  #> 0

  a.max()
  #> 8
  ```

---

## Functions that work element-wise

- Square root:

  ```python
  np.sqrt(a)
  #> array([0., 1., 1.41421356, 1.73205081, 2., 2.23606798,
  #>        2.44948974, 2.64575131, 2.82842712])
```

- Exponentiation:

  ```python
  np.exp(a)
  #> array([1.00000000e+00, 2.71828183e+00, 7.38905610e+00,
  #>        2.00855369e+01, 5.45981500e+01, 1.48413159e+02,
  #>        4.03428793e+02, 1.09663316e+03, 2.98095799e+03])
  ```

---

## Creating arrays by converting lists

Convert a list to a one-dimensional array:

- If we create an array of numbers without digits,
  they will be interpreted as integers:

  ```python
  a1 = np.array([1, 2, 3, 4])
  print(a1)
  #> [1 2 3 4]
  
  a1.dtype.name
  #> 'int64'
  ```

- If we include decimal points, they will be floats:

  ```python
  a1 = np.array([1.0, 2.0, 3.0, 4.0])
  
  print(a1)
  #> [ 1. 2. 3. 4.]
  a1.dtype.name
  #> 'float64'
  ```

---

## Creating arrays by converting lists (cont.)

Convert a list of lists to a two-dimensional array.

- Integers:

  ```python
  m = np.array([[1, 2], [3, 4]])
  m
  #> array([[1, 2],
  #>       [3, 4]])
  ```
  
- Floats - now we explicitly specify the data type:  
  
  ```python
  m = np.array([[1, 2], [3, 4]], dtype = float)
  
  print(m)
  #> [[ 1. 2.]
  #> [ 3. 4.]]
  m.dtype.name
  #> 'float64'
  ```

---

## Initialize arrays with a given structure

- `np.zeros()` will create an array with 0s.

  We can specify the dimensions with a tuple &ndash;
  here, we ask for a 3x2 array filled with floating-point zeros:

  ```python
  m = np.zeros((3, 2), dtype = float)
  m
  #> array([[ 0., 0.],
  #> [ 0., 0.],
  #> [ 0., 0.]])
  ```

--

- We can also "reshape" an existing array using `reshape()` &ndash;
  here, we create an 1-dimensional array containing the integers from 0 to 8,
  and then arrange it in a 3x3 array:
  
  ```python
  a = np.arange(9)
  a.reshape((3,3))
  #> array([[0, 1, 2],
  #>       [3, 4, 5],
  #>       [6, 7, 8]])
  ```

---

## Initialize arrays with a given structure (cont.)

- NumPy can also generate random numbers with the `np.random` module (more later).

  Let's create an array with random values,
  drawn from the uniform distribution `U[0,1]`
  (the default for `np.random.random`):

  ```python
  np.random.random((2, 3))
  #> array([[ 0.2331427 , 0.28167952, 0.66094357],
  #>       [0.13703488, 0.75519455, 0.08413554]])
  ```
  
  Note that with the tuple `(2, 3)`, we merely specify the dimensions.
  
---

## Array indexing

With one-dimensional arrays, indexing works just like we have seen for lists.

```python
a
#> array([0, 1, 2, 3, 4, 5, 6, 7, 8])
```
  
- Single-element indexing &ndash; e.g., get the 2nd element:
  
  ```python
  a[1]
  #> 1
  ```

- Create a slice &ndash; e.g., from the start to the 5th element (noninclusive):
  
  ```python
  a[:4]
  #> array([0, 1, 2, 3])
  ```
  
- Slice with negative indices &ndash; e.g., from the 3rd-to-last to the last
  element:
  
  ```python
  a[-3:]
  #> array([6, 7, 8])
  ```
---

## Two-dimensional indexing

```python
m = np.arange(16).reshape((4, 4))
m
#> array([[ 0, 1, 2, 3],
#>        [ 4, 5, 6, 7],
#>        [ 8, 9, 10, 11],
#>        [12, 13, 14, 15]])
```

- Extract a submatrix &ndash; e.g. rows 1-3 and columns 2-4:
  
  ```python
  m[0:3, 1:4]
  #> array([[ 1, 2, 3],
  #>        [ 5, 6, 7],
  #>        [ 9, 10, 11]])
  ```
  
.pull-left[
- Get the second row:

  ```python
  m[1, :]
  #> array([4, 5, 6, 7])
  m[1]
  #> array([4, 5, 6, 7])
  ```
]

.pull-right[  
- Get the second column:

  ```python
  m[:, 1]
  #> array([ 1, 5, 9, 13])
  ```
]

---

## Operate on the whole array, rows, or columns

- By default, methods like `sum()` will operate on the entire matrix:

  ```python
  m.sum()
  #> 120
  ```
  
<br>

- To get sums for each **column**, we specify **`axis = 0`**:

  ```python
  m.sum(axis = 0)
  #> array([24, 28, 32, 36])
  ```

- To get sums for each **row**, we specify **`axis = 1`**:
  
  ```python
  m.sum(axis = 1)
  #> array([ 6, 22, 38, 54])
  ```

---

## Load data into NumPy

Rather than creating *file handles* and reading data often line-by-line,
as we did in the past few weeks,
NumPy (as well as Pandas later on) typically
**read files into memory all at once as an object**.

For instance, we can use NumPy's `loadtxt()` function as follows:

```python
m = np.loadtxt("my_matrix_file.csv", delimiter = ",")
```

---

## Example: image processing with NumPy


---

## Random numbers and distributions (6.2.2)

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>


