---
title: "Week 11 - Python: Scientific computing"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 11 - Python: Scientific computing*

----

# II: Pandas

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/25 (updated: `r Sys.Date()`)

---

## Pandas

Pandas is short for "Python Data Analysis Library".

As NumPy has arrays, the signature data structure of Pandas it its
`DataFrame`.
**With Pandas' `DataFrame`s, we can analyze tabular data much more
smoothly in Python.**

--

If you are familiar with R's *tidyverse*,
Pandas allows for similar types of operations on tabular data:

- Basic manipulations such as removing and adding columns, filtering data.

- Merging (joining) data sets based on shared columns.

- Summarize data including with "split-apply-combine":
  compute statistics for multiple groups at once.

- Pivoting from "wide" to "long" data.

- Integration with a plotting library, Matplotlib.

---

## Getting set up

Panda is built on NumPy, so one often imports NumPy along with Pandas:

```python
import pandas as pd
import numpy as np
```

Next, let's move into the CSB chapter's sandbox directory:

```python
import os
os.chdir('/fs/ess/PAS1855/users/<user>/CSB/scientific/sandbox')
# os.chdir('../CSB/scientific/sandbox')
```

---

## Reading in tabular data

For tabular data to be analyzed by NumPy and Pandas,
we don't typically read files line-by-line,
but we load them all at once &ndash;  
the standard Pandas function for this is **`read_csv()`**:

```python
data = pd.read_csv("../data/Dale2015_data.csv")
```

<br>

.content-box-info[
The argument `sep` can be used to specify the delimiter, e.g. for tab-delimited
files (`sep = '\t'`), though Pandas can typically auto-detect this too.

There are corresponding functions to read in **Excel** and **JSON** files,
among others.
]

---

## Exploring DataFrames

- Show the first 6 rows with `head()`:

  ```python
  data.head()
  
  #>      Scientific_name	         English_name	          TipLabel	Female_plumage_score	Male_plumage_score
  #> 0	Abroscopus albogularis	  Rufous-faced Warbler	  Abroscopus_albogularis	45.833333	51.944444
  #> 1	Abroscopus schisticeps	  Black-faced Warbler	    Abroscopus_schisticeps	48.888889	47.777778
  #> 2	Abroscopus superciliaris	Yellow-bellied Warbler	  Abroscopus_superciliaris	43.194444	43.611111
  #> 3	Acanthagenys rufogularis	Spiny-cheeked Honeyeater	Acanthagenys_rufogularis	39.305556	39.166667
  #> 4	Acanthidops bairdii	     Peg-billed Finch	        Acanthidops_bairdii	40.555556	54.166667
  ```

- Just typing the name of the dataframe will also return a very useful view
  in our interactive Python window:

  ```python
  data
  ```

---

## Exploring DataFrames: attributes

- **Note that NumPy functions, methods, and attributes will also work for
  Pandas `DataFrames`!**
  
  For instance, we can check the length along each dimension with `shape`:

  ```python
  data.shape
  #> (5831, 5)
  ```

---

## Exploring DataFrames: data types

- `dtypes` returns data types by column (note, strings are called `object`):

  ```python
  data.dtypes
  #> data_raw.dtypes
  #> Scientific_name          object
  #> English_name             object
  #> TipLabel                 object
  #> Female_plumage_score    float64
  #> Male_plumage_score      float64
  #> dtype: object
  ```

.content-box-info[
Pandas `DataFrame`s can be thought of as
**a set of *aligned* 1-dimensional NumPy arrays of equal length**,
with each `DataFrame` column representing an array.

Each column contains elements of the same data type while different columns
typically have different data types.

(You can also think of `DataFrame`s is as a **special kind of dictionary**:
each column is an item with column names as keys.)
]

---

## Exploring DataFrames: indices

- `columns` returns an `Index` object with the column names:

  ```python
  data.columns
  #> Index(['Scientific_name', 'English_name', 'TipLabel',
  #>        'Female_plumage_score', 'Male_plumage_score'],
  #>        dtype='object')
  ```

<br>

- A `DataFrame` also has an explicit **rowwise index**  
  (noticed the row numbering when we printed `data`?):
  
  ```python
  data.index
  #> RangeIndex(start=0, stop=5831, step=1)
  ```
  
  This index can also be manipulated and could for example consist of *strings*
  as well.

---

## Exploring DataFrames: describe()

The method `describe()` gives an overview of the numerical columns:

```python
data.describe()

#> 	     Female_plumage_score	  Male_plumage_score
#> count	        5831.000000	         5831.000000
#> mean	           47.527630	           51.009189
#> std	             6.997343	            8.200663
#> min	            34.166667	           34.027778
#> 25%	            42.361111	           44.166667
#> 50%	            45.833333	           49.722222
#> 75%	            51.250000	           57.083333
#> max	            72.222222	           74.166667
```

---

## Exploring DataFrames: info()

And `info()` gives another type of useful overview, for all columns:

```python
data.info()

#> <class 'pandas.core.frame.DataFrame'>
#> RangeIndex: 5831 entries, 0 to 5830
#> Data columns (total 5 columns):
#>  #   Column                Non-Null Count  Dtype  
#> ---  ------                --------------  -----  
#>  0   Scientific_name       5831 non-null   object 
#>  1   English_name          5831 non-null   object 
#>  2   TipLabel              5831 non-null   object 
#>  3   Female_plumage_score  5831 non-null   float64
#>  4   Male_plumage_score    5831 non-null   float64
#> dtypes: float64(2), object(3)
#> memory usage: 227.9+ KB
```

---

## Exploring DataFrames: selecting columns

We can specify columns not with the syntax `df["colname"]` (think dictionary!), as well as with the shorthand `df.colname`:
  
```python
data["Scientific_name"]

data.Scientific_name

#> 0         Abroscopus albogularis
#> 1         Abroscopus schisticeps
#> 2       Abroscopus superciliaris
#> 3       Acanthagenys rufogularis
#> 4            Acanthidops bairdii
#>                   ...           
#> 5826       Zosterops uropygialis
#> 5827          Zosterops vaughani
#> 5828      Zosterops vellalavella
#> 5829          Zosterops wallacei
#> 5830       Zosterops xanthochroa
#> Name: latin, Length: 5831, dtype: object
```
  
When we extract a single column,
the resulting 1-dimensional Pandas array is called a `Series`.
  
---

## Manipulating DataFrames: renaming columns

- We can rename columns using `rename()` with a
  **dictionary of old and new column names.**
  
  Let's do so to make the column names and therefore our lines of code
  a bit shorter:
  
  ```python
  data = data.rename(columns={"Scientific_name": "latin",
                                "English_name": "english",
                                "Female_plumage_score": "score_f",
                                "Male_plumage_score": "score_m"})
  ```

---

## Manipulating DataFrames: adding columns

- Create a new column that sums the plumage score for males and females:

  ```python
  data["score_sum"] = data["score_f"] + data["score_m"]
  
  data.head()
  ```

--

<br>

- We could also easily add a column with a constant using a vectorized
  operation, like we have seen with NumPy:
  
  ```python
  data["Study"] = 1
  
  data.head()
  ```

---

## Manipulating DataFrames: removing columns

- We can use the familiar, generic `del()` function to drop one column,  
  or the more verbose `drop()` method to remove one or more columns:
  
  ```python
  del(data["score_sum"])
  
  data.drop(["TipLabel", "Study"], axis = 1, inplace = True)
  ```
  
  - `axis = 0` acts along rows, and `axis = 1` along columns.
  
  - `inplace` means that the `DataFrame` is modified in place.


---

## Thursday setup

```python
import pandas as pd
import numpy as np
import os

username = os.environ.get('USER')
os.chdir('/fs/ess/PAS1855/users/' + username + '/CSB/scientific/sandbox')
```

```python
data = pd.read_csv("../data/Dale2015_data.csv")

data = data.rename(columns={"Scientific_name": "latin",
                            "English_name": "english",
                            "Female_plumage_score": "score_f",
                            "Male_plumage_score": "score_m"})
```

---

## Indexing and slicing DataFrames

- After selection a column, we can slice this the way we may expect to
  for an array/list:
  
  ```python
  data["latin"][:3]                  # First three  
  #> 0 Abroscopus albogularis
  #> 1 Abroscopus schisticeps
  #> 2 Abroscopus superciliaris
  ```

<br>

- But we can't slice the entire `DataFrame` as we may have expected
  after learning NumPy:

  ```python
  data[:3, :3]
  #> TypeError: '(slice(None, 3, None), slice(None, 3, None))' is an invalid key
  ```

---

## Indexing and slicing DataFrames (cont.)

- Instead, we need to use the **`iloc()` method**:
  
  ```python
  data.iloc[2, 1]
  #> 'Yellow-bellied Warbler'
  
  # First 2 rows, all columns (out-of-range slices work):
  data.iloc[:2, :10]
  
  # First 2 rows, columns 2 and 4:
  data.iloc[:2, [1, 3]]
  
  # Last 5 rows, all columns:
  data.iloc[-5:, :]
  ```
  
---

## Indexing and slicing DataFrames (cont.)

We can also use the **`loc` method**,
which selects rows and columns using **explicit labels**.  

The labels for the columns are &ndash;of course&ndash; the column names,  
whereas the labels for the rows are in the **row number index**.

Notice that ranges given with `loc` are *inclusive*:  
so `data.loc[:3, ]` returns the first *four* rows!

```python
data.loc[:3, ["latin", "english"]]
#> 0 Abroscopus albogularis   Rufous-faced Warbler
#> 1 Abroscopus schisticeps   Black-faced Warbler
#> 2 Abroscopus superciliaris Yellow-bellied Warbler
#> 3 Acanthagenys rufogularis Spiny-cheeked Honeyeater
```

---

## Masking (filtering) DataFrames

Like we've seen for NumPy arrays, we can also perform **masking**:  
filter the data based on (a `Series` of) Booleans.

- For instance, to select the row(s) for a certain species:

  ```python
  data["latin"] == "Zosterops mouroniensis"
  #> 0       False
  #> 1       False
  #> 2       False
  #> ...
  
  data.loc[data["latin"] == "Zosterops mouroniensis"]
  #         Scientific_name	       English_name	            TipLabel	Female_plumage_score	Male_plumage_score	Sum_scores	Study
  # 5801	Zosterops mouroniensis	Mount Karthala White-eye	Zosterops_mouroniensis	47.916667	47.5	95.416667	1
  ```

--

- Or get only the most highly dimorphic species:

  ```python
  data[(data["score_m"] > 70) & (data["score_f"] < 40)]
  #> 	      latin	              english	TipLabel	      score_f	score_m
  #> 875	   Chiroxiphia pareola	Blue-backed Manakin	     Chiroxiphia_pareola	39.305556	71.666667
  #> 1355	  Cyornis hainanus	   Hainan Blue-flycatcher	  Cyornis_hainanus	37.222222	71.250000
  #> 1811	  Euplectes hartlaubi	Marsh Widowbird	         Euplectes_hartlaubi	39.583333	70.555556
  ```

---

<figure>
<p align="center">
<img src=img/Blue-backedManakin.jpeg width="62%">
<figcaption>Blue-backed Manakin - males (<a href="http://www.arctracer.com/photos/trinidad/birds/Blue-backedManakin.html">Figure source</a>)</figcaption>
</p>
</figure>

<figure>
<p align="center">
<img src=img/manakin_female.jpg width="62%">
<figcaption>Blue-backed Manakin - female (<a href="http://www.arthurgrosset.com/sabirds/photos/chipar28910.jpg">Figure source</a>)</figcaption>
</p>
</figure>

---

<figure>
<p align="center">
<img src=img/marsh_widowbird.jpg width="60%">
<figcaption>Marsh Widowbird - male (<a href="http://www.arthurgrosset.com/sabirds/photos/chipar28910.jpg">Figure source</a>)</figcaption>
</p>
</figure>

<figure>
<p align="center">
<img src=img/marsh_widowbird_female.jpg width="60%">
<figcaption>Marsh Widowbird - female (<a href="http://www.arthurgrosset.com/sabirds/photos/chipar28910.jpg">Figure source</a>)</figcaption>
</p>
</figure>

---

## Summary statistics

Like we have seen with NumPy, we can also easily get summary statistics:

```python
data.score_m.mean()
#> 51.009189390042877

data.score_m.median()
#> 49.72222222

data.score_m.std()
#> 8.2006629346736908
```

---

## Some simple plots with Matplotlib &ndash; histogram:

```python
data.score_m.hist()
```

<p align="center">
<img src=img/hist.svg width="90%">
</p>

---

## Simple plots with Matplotlib &ndash; scatterplot

```python
data.plot.scatter(x = "score_m", y = "score_f")
```

<p align="center">
<img src=img/scatter.svg width="90%">
</p>

---

## Simple plots with Matplotlib &ndash; boxplot:

```python
data[["score_m", "score_f"]].plot.box()
```

<p align="center">
<img src=img/boxplot.svg width="90%">
</p>

---

## <i class="fa fa-user-edit"></i> Your turn

1. Retrieve the English names of species for which both males and females have a
   plumage score higher than 70.
   
2. For how many species do either males, females, or both have plumage scores
   lower than 50?

---

## <i class="fa fa-user-edit"></i> Your turn: Solutions

1\. Retrieve the English names of species for which both males and females have a
    plumage score higher than 70.
   
```python
data[(data["score_m"] > 70) & (data["score_f"] > 70)].english
#> 363            Unicoloured Jay
#> 1520      Indigo Flowerpiercer
#> 1717               Red Warbler
#> 1736    Red-headed Parrotfinch
#> 2739        Red-headed Malimbe
#> Name: english, dtype: object
```

--

<br>

2\. For how many species do either males, females, or both have plumage scores
    lower than 50?

--

```python
data[(data["score_m"] < 50) | (data["score_f"] < 50)].shape[0]
#> 4246
```

---

## <i class="fa fa-user-edit"></i> Pandas exercise for homework

The third exercise for this week uses Pandas to investigate whether
scientific manuscript rejection rates are higher when the manuscript gets
more reviewers.

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Bonus material

-----

<br> <br> <br> <br>

---
background-color: #f2f5eb

## "Views" versus copies in Pandas

When we slice and filter `DataFrames`, we generally get a "**view**" of the
data, as opposed to a *copy* of the data.
Therefore, we get a warning when we try to modify such as view:

```python
high_male_score = data[data["score_m"] > 65]

high_male_score["score_qual"] = "High"
#> [...] SettingWithCopyWarning:
#> A value is trying to be set on a copy of a slice from a DataFrame.
```

--

Our operation still worked &ndash;
a column has been added to `high_male_score` but not to data:

```python
high_male_score.columns
#>Index(['latin', 'english', 'TipLabel', 'score_f', 'score_m', 'score_qual'],
#>      dtype='object')

data.columns
#> Index(['latin', 'english', 'TipLabel', 'score_f', 'score_m'],
#>       dtype='object')
```

---
background-color: #f2f5eb

## "Views" versus copies in Pandas (cont.)

But in such case, we should instead make a copy of the data:

```python
high_male_score = data[data["score_m"] > 65].copy()

high_male_score["score_qual"] = "High"
```

---
background-color: #f2f5eb

## Masking (filtering) DataFrames

- We can even select rows based on only part of the cell content,  
  using the **`str.contains()`** method:

  ```python
  data[data.english.str.contains("Bower")]
  #> [All rows containing "Bower" in the "english" column]
  
  data[data.english.str.contains("Bower")]["latin"][:3]
  #> 188 Amblyornis flavifrons
  #> 189 Amblyornis inornata
  #> 190 Amblyornis macgregoriae
  #> Name: Scientific_name, dtype: object
  ```

.content-box-q[
Try this:

```python
data[data.english.str.contains("Bower")].loc[:2, "latin"]
```

Why is it not returning any rows?
]
