---
title: "Week 13 - Reproducible workflows with Snakemake"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)

icon::fa("user-edit")
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
xaringanExtra::use_scribble()
xaringanExtra::use_panelset()
xaringanExtra::use_clipboard()
xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 13 &ndash; <br> Reproducible workflows with Snakemake*

----

# I: Workflows and workflow tools

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/04/06 (updated: `r Sys.Date()`)

---

## Overview of this week

- I: Intro to workflows and workflow tools &mdash; today
  
- II: Getting started with Snakemake &mdash; today/Thu

- III: More advanced Snakemake tips and tricks &mdash; Thu

<br>

Among other things, the **exercises** will walk you through some more Snakemake
aspects such as running Snakemake on a cluster.

---

## What do we mean by a "pipeline" or "workflow"?

<br>

<p align="center">
<img src=img/schem-workflow.svg width=75%>
</p>

---

## An example partial workflow script

```sh
samples=(smpA smpC smpG)

# Trim:
for sample in ${samples[@]}; do
    scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
done

# Map:
for sample in ${samples[@]}; do
    scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
done

# Count:
scripts/count.sh ${samples[@]} > res/count_table.txt
```

---

## An example partial workflow script

With a single `for` loop:

```sh
samples=(smpA smpC smpG)

# Trim and map:
for sample in ${samples[@]}; do
    scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
    scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
done

# Count:
scripts/count.sh ${samples[@]} > res/count_table.txt
```

---

## Why create a pipeline / workflow?

What are the advantages of creating such a pipeline or workflow,
rather than running scripts one-by-one as needed?

<br>

- **Rerunning** everything is much easier &ndash; you may want to do so after
  adding samples, changing a script, changing settings, and so on.

- **Re-applying** the same set of analyses in a different project is much easier. 

<br>

--

- The workflow is a form of **documentation** of the steps taken.

- It makes sure you are **including all necessary steps** and
  are not overlooking a script or a manual step.

- **Reproducibility** in is generally improved,
  e.g. it will be easier for others to repeat you analysis.

---

## Challenges with workflows: <br> You need to rerun parts of it

**Rerunning *part* of the workflow may be necessary**, e.g. after:
  
- Some scripts fail.

- Some scripts fail for some samples only.

- Adding a sample.

- You want to change a script or settings somewhere halfway the workflow.

---

## Challenges with workflows: <br> You need to rerun parts of it

**Ad-hoc strategies to solve some of these:**
  
- Comment out part of the workflow &ndash; to skip a step:

  ```sh
  for sample in ${samples[@]}; do
        #scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
        scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
  done
  scripts/count.sh ${samples[@]} > res/count_table.txt
  ```

--

- Make temporary changes &ndash; to run a single new sample:

  ```sh
  samples=(smpA smpC smpG smpT)
  
  #for sample in ${samples[@]}; do
        sample=smpD  # TEMPORARY LINE
        scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
        scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
  #done
  scripts/count.sh ${samples[@]} > res/count_table.txt
  ```

---

## Challenges with workflows: <br> You need to rerun parts of it

**Labor-intensive strategy to solve some of these:**  
Command-line options and `if`-statements to enable flexible running of
part of the workflow (and perhaps to change settings):

```sh
trim=$1   # true or false
map=$2    # true or false
count=$3  # true or false

for sample in ${samples[@]}; do
    if [ "$trim" = true ]; then
        scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
    fi
    
    if [ "$map" = true ]; then
        scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
    fi
done

if [ "$count" = true ]; then
    scripts/count.sh ${samples[@]} > res/count_table.txt
fi
```

---

## Challenges with workflows: <br> managing cluster jobs

Say we want to speed things up and submit SLURM jobs for each script:

```sh
for sample in ${samples[@]}; do
    sbatch scripts/trim.sh data/"$sample".fq > res/"$sample"_trim.fq
    sbatch scripts/map.sh res/"$sample"_trim.fq > res/"$sample".bam
done

sbatch scripts/count.sh ${samples[@]} > res/count_table.txt
```

.content-box-q[
What is the problem here?
]

--

.content-box-answer[
Steps that need output files from prior steps won't wait for those prior steps!
]

For any workflow that has some steps operating on each sample separately:
**How do you let the steps properly wait for each other?**


---

## Challenges with workflows: <br> managing cluster jobs

**Ad-hoc strategies:**

- Go back to the original script and **run the whole workflow as a single job,**
  in which per-sample steps are simply run 1-by-1 in a loop.  
  This will be too time-consuming in many cases...

- Rather than running the workflow script as an actual script,  
  just **use it as scaffolding** and submit jobs one-by-one / group-by-group. 
  
--

**Labor-intensive strategy:**

- Use SLURM job dependencies &ndash;
  but hard to manage for more complex workflows.
  
  ```sh
  for sample in ${samples[@]}; do
      line=$(sbatch scripts/trim.sh)
      trim_id=$(echo $line | sed 's/Submitted batch job //')
      sbatch --dependency=afterok:$trim_id scripts/map.sh
  done
  
  sbatch --dependency=singleton scripts/count.sh
  ```

---

## Challenges with workflows: <br> monitor and manage script failure

**How do you monitor job failure and automatically stop the workflow upon job
failure?**

<br>

**Ad-hoc strategies (same as earlier):**

- Go back to the original script and **run the whole workflow as a single job**:
  proper Bash settings will at least make the workflow stop upon failure.

- Rather than running the workflow script as an actual script,  
  just **use it as scaffolding** and submit jobs one-by-one / group-by-group.
  
<br>

--

**Labor-intensive strategy:**

- Lots of testing for proper input for and exit status of each script, etc.

---

## The need for specialized workflow tools

<p align="center">
<img src=img/nature-feature.png width=100%>
</p>

----

--

> Typically, researchers codify workflows using general scripting languages such as Python or Bash. But these often lack the necessary flexibility.

>Workflows can involve hundreds to thousands of data files; a pipeline must be able to monitor their progress and exit gracefully if any step fails. And pipelines must be smart enough to work out which tasks need to be re-executed and which do not.

---

## The need for specialized workflow tools?

<p align="center">
<img src=img/nature-feature.png width=100%>
</p>

<br>

> Bioinformatician Davis McCarthy at St Vincent’s Institute of Medical Research in Fitzroy, Australia, says Python and R were more than enough for the relatively simple workflows he used as a PhD student.

---

## Workflow management systems

Workflow tools, often called "workflow management systems",  
provide ways to **formally describe and execute workflows.**

<br>

Some of the most commonly used command-line based options for working with
biological sequence data are:

- Snakemake
- Nextflow
- Common Workflow Language (CWL) 
- Workflow Description Language (WDL)

<br>

--

While there is a bit of a learning curve for all of these,  
they solve the challenges discussed above, and offer several other benefits.

For many workflows, they will eventually be huge time-savers,
both in terms of time spent composing the workflow and in running time of the
workflow.

---

## Advantages of formal workflow management &ndash; <br> all the buzzwords...

- **Reproducibility and transparency**
  - Document dependencies among data and among code.
  - Easy visualization of a workflow graph.
  - Integration with software management.

- **Automation**

  - Detect & rerun upon changes in input files and failed steps.
  - Rerun for other datasets.
  - Automate SLURM job submissions.

- **Portability**
  
  - Workflow can be easily rerun on a laptop, an HPC system,
    in the cloud, with or without containers.
    
---

## Advantages of formal workflow management (cont.)

- **Scalability**

  - Efficient execution &ndash; all dependencies are known so cluster submission
    can be optimized.
  

<br>

- **Flexibility** due to a separation of:
  
  - Generic nuts-and-bolts of the workflow.
  - Run-specific configuration &ndash; samples, directories, options.
  - Things specific to the runtime environment (laptop vs cluster vs cloud).

<br>

--

- **Convenience**
  
  - Reduces boilerplate code: no need to loop over samples yourself, etc.
  - Some of the points mentioned above such as automated job submission and
    integrated software management.
  
---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---

<figure>
<p align="center">
<img src=img/giaa140fig1.jpeg width="90%">
<figcaption>From <a href="https://academic.oup.com/gigascience/article/10/1/giaa140/6092773/">Reiter et al. 2021: Streamlining data-intensive biology with workflow systems</a></figcaption>
</p>
</figure>

---

## Advantages of formal workflow management

> The essence of what we're trying to do:
>
> - Automate a workflow
> - Document the workflow
> - Document the dependencies among data files, code
> - Re-run only the necessary code, based on what has changed
>
> &mdash; [Karl Broman](http://kbroman.org/Tools4RR/assets/lectures/01_intro.pdf)
