---
title: "Week 10 - Writing Good Code"
output:
  xaringan::moon_reader:
    seal: false
    css: ["default", "default-fonts", "slides.css", "slides_copy.css"]
    lib_dir: libs
    nature:
      highlightStyle: rainbow
      highlightLines: true
      countIncrementalSlides: false
---
class:inverse middle center

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

```{r xaringan-extra, echo=FALSE, eval=TRUE}
#xaringanExtra::use_scribble()
#xaringanExtra::use_panelset()
#xaringanExtra::use_clipboard()
#xaringanExtra::use_search(show_icon = TRUE)
```

# *Week 10 - Writing good code*

----

# II: Calling Python scripts, errors, <br> and beyond the basics

<br> <br> <br> <br> <br>

### Jelmer Poelstra
### 2021/03/18 (updated: `r Sys.Date()`)

---
class: center middle inverse

# Overview

----

### - Calling Python scripts (4.4)

### - Beyond the basics (4.9)

### - Errors and error handling (4.5)

---
class: center middle inverse

# Calling Python scripts (4.4)

-----

<br> <br> <br> <br>

---

## Calling Python scripts

So far, we have worked with Python:
  
  - By typing in the Python interactive window or console directly.
  
  - Similarly, by interactively sending code from a script there.
  
  - By loading a Python script as a module using `import my_module`.

But, like we did with Bash scripts, we can also save a script and then  
**run it non-interactively by calling it from the shell.**

--

<br>

----

<br>

**We will practice this with our `simulate_drift.py` script from Tuesday.**

---

## Calling Python scripts (cont.)

To be able to run the script smoothly and flexibly, we will:

- Add a *shebang* line (inside the script).

- Make the script executable (outside of the script).

- Parse command-line arguments passed to the script  
  (inside & outside of the script).

- Add an `if __name__ == "__main__"` statement (inside the script).

<br>

Most of this is familiar from what we did running Bash scripts!

---

## Calling Python scripts: *shebang* line

We could write a *shebang* analogous to what we did for Bash scripts,
providing `/usr/bin/python3` (vs. `/bin/bash`) as the path to the interpreter.
This is also what CSB does.

However, we will instead **call the `env` utility to determine the path of Python
for us.**

--

Let's see what difference that makes &ndash; open a new terminal and type:

```sh
$ /usr/bin/env python3
#> ...
#> Python 3.6.8
>>>                     # Ctrl-D or exit() to exit Python

$ which python3
#> /usr/bin/python3
```

That loaded OSC's default Python 3 installation at `/usr/bin/python3`.

---

## Calling Python scripts: *shebang* line

If we now load our Conda environment with a new version of Python,  
and try again:

```sh
$ module load python/3.6-conda5.2
$ source activate ipy-env
$ /usr/bin/env python3
#> ...
#> Python 3.9.2
>>>                     # Ctrl-D or exit() to exit Python

$ which python3         # Check the current path to Python
#> ~/.conda/envs/ipy-env/bin/python3
```

Nice! `env` automatically tracked down the correct version of Python for us.

So, using `/usr/bin/env python3` in the *shebang* line is flexible,
and we don't need to change the path depending on which Python installation
we want to use.

---

## Calling Python scripts: Make the script executable

Make the script executable in the shell &ndash;
same as we've done with Bash scripts:

```sh
$ chmod +x simulate_drift.py
```

---

## Calling Python scripts: Parse command-line arguments

Again like in our Bash scripts,
we also generally want scripts to take *arguments*.

Arguments that were passed to the script (as in `./my.py arg1 arg2`) are
available inside the script in the list `sys.argv`.

Just like `$0` contained the name of the script in Bash,
and `$1` contained the first argument, and so on, in Python:

- `sys.argv[0]` contains the name of the script.
- `sys.argv[1]` contains the first argument.
- `sys.argv[1]` contains the second argument.
- And so on...

---

## Calling Python scripts: Parse command-line arguments

To parse the command-line arguments, we therefore include the following code
in the script:

```python
# DON'T COPY THIS YET, WE WILL MAKE MODIFICATIONS

# Parse command line arguments, and convert to the right type
# (they are strings by default):
import sys ?

N = int(sys.argv[1])
p = float(sys.argv[2])
```

<br>

Additionally, our script currently just contains a function.
We therefore also need to *call* the function in order to run it,
and we use the command-line arguments to do so:

```python
# Call the simulation:
simulate_drift(N, p)
```

---

## Calling Python scripts: `__name__`

A Python script can be run in two quite different ways:

- It can be imported into any Python session using `import`.
  When we do this, we typically just want to load the functions that are 
  available in the script.

- They can be run from the shell.
  In this case, we want to actually run some computations,
  e.g. *call* the functions in the script. 

---

## Calling Python scripts: `__name__`

We could create separate scripts for both use cases, but Python makes it easy
for a single script to perform either of these functions depending on the context.

Using the special variable `__name__`, we can tell the use case:

- If the script is being imported,  
  **the value of `__name__` will be the name of the script.**

- If the script is being called from the shell,  
  **the value of `__name__` will be `'__main__'`.**
  
---

## Calling Python scripts: `__name__`

Therefore, we can include the following `if` statement to determine whether
a script is being called from the shell &mdash; and *only* if it is, we actually
parse arguments and *run* the functions in our script:

```python
if __name__ == "__main__":

  # Parse command line arguments, and convert to the right type
  # (they are strings by default):
  N = int(sys.argv[1])
  p = float(sys.argv[2])
  
  # Call the simulation:
  simulate_drift(N, p)
```

--

While the `__name__ == "__main__"` call is not necessary *just* for our
current use case
(we want to call the script to run it and may not have an intention to import it),
it is common and good practice for Python scripts to include this.

That way, our scripts can always be called directly *or* loaded as a module
without problems.

---

## Finally, calling our script

Now, we can run our Python script from the shell:

```sh
$ ./simulate_drift.py 1000 0.1
```

---
class: center middle inverse

# Beyond the basics (4.9)

-----

<br> <br> <br> <br>

---

## Arithmetic of data structures

- Lists:

  ```python
  a = [1, 2, 3]
  b = [4, 5]
  
  a + b
  ```

---

## Arithmetic of data structures

- Lists:

  ```python
  a = [1, 2, 3]
  b = [4, 5]
  
  a + b
  #> [1, 2, 3, 4, 5]
  ```

--

- Tuples:

  ```python
  a = (1, 2)
  b = (4, 6)
  a + b
  #> (1, 2, 4, 6)
  ```

---

## Arithmetic of data structures (cont.)

- Sets:

  ```python
  a = {1, 2}
  b = {4, 6}
  a + b
  #> TypeError: unsupported operand type(s) for +: 'set' and 'set'
  ```

---

## Arithmetic of data structures (cont.)

- Sets:

  ```python
  a = {1, 2}
  b = {4, 6}
  a + b
  #> TypeError: unsupported operand type(s) for +: 'set' and 'set'
  ```

---

## Arithmetic of data structures (cont.)

- Sets:

  ```python
  a = {1, 2}
  b = {4, 6}
  a + b
  #> TypeError: unsupported operand type(s) for +: 'set' and 'set'  
  ```
  
  ```python
  a | b
  {1, 2, 4, 6}
  ```

--

- Dictionaries:
  
  ```python
  z1 = {1: "AAA", 2: "BBB"}
  z2 = {3: "CCC", 4: "DDD"}
  z1 + z2
  #> TypeError: unsupported operand type(s) for +: 'dict' and 'dict'
  ```
  
--

  ```python
  z1.update(z2)
  z1
  #> {1: 'AAA', 2: 'BBB', 3: 'CCC', 4: 'DDD'}
  ```

---

## Arithmetic of data structures (cont.)

We can also "multiply" strings, lists, and tuples,
though recall that this is never element-wise:

```python
'a' * 3
#> 'aaa'

(1, 2, 4) * 2
#> (1, 2, 4, 1, 2, 4)

[1, 3, 5] * 4
#> [1, 3, 5, 1, 3, 5, 1, 3, 5, 1, 3, 5]
```

---

## Mutable and immutable data types

*Strings*, *sets*, and *tuples* are **immutable**. As a consequence:

- They don't have methods that can modify them *in place*,
  and any "modification" that you may seem to be making creates a completely
  new object.
  
  ```python
  seq[2] = "U"
  #> TypeError: 'str' object does not support item assignment
  ```

--

- Assigning a new variable name will always create an entirely new,
  unlinked variable:

  ```python
  x = 5
  y = x
  y = y + 10
  y
  #> 15
  x
  #> 5
  ```

---

## Mutable and immutable data types

We have seen that *lists*, *dictionaries*, and *sets* are **mutable**.
As a consequence:

- They have methods that can modify them *in place*.

  ```python
  seq = list('AACG')
  reversed_seq_maybe = seq.reverse()
  
  reversed_seq_maybe
  #>
  print(reversed_seq_maybe)
  #> None
  
  seq
  #> ['A', 'A', 'C', 'G']
  ```

- When we assign them to a new variable name, all that happens is that we create
  a new pointer to the *same variable*, and if we make a change using one of the
  names, the other is changed too.

---

## Mutable and immutable data types

We have seen that *lists*, *dictionaries*, and *sets* are **mutable**.
As a consequence:

- They have methods that can modify them *in place*.

- When we assign them to a new variable name, all that happens is that we create
  a new pointer to the *same variable*, and if we make a change using one of the
  names, the other is changed too.
  
  ```python
  seq2 = seq
  
  seq2 = seq2.append('T')
  
  seq2
  #> ['A', 'A', 'C', 'G', 'T']
  seq
  #> ['A', 'A', 'C', 'G', 'T']
  ```

  To actually create a new object with the same contents, we need to explicitly
  make a copy.

---

## Copying mutable objects 

- The `copy()` method will create a copy:

  ```python
  import copy
  seq2 = seq.copy()
  seq
  #> ['A', 'A', 'C', 'G', 'T']
  ```

--

- Now, these lists can be modified independently:

  ```python
  seq.clear()
  
  seq
  #> []
  
  seq2
  #> ['A', 'A', 'C', 'G', 'T']
  ```

--

.content-box-info[
For lists specifically, we can also use `[:]` notation:

```python
seq2 = seq[:] 
```
]

---

## Copying mutable objects 

**However**, for "nested" objects, such as lists containing lists or dictionaries
containing dictionaries, the "shallow" copy made by `copy` does not suffice:

```python
a = [[1, 2], [3, 4]]
b = a.copy()

a[0][1] = 10

a
#> [[1, 10], [3, 4]]
b
#> [[1, 10], [3, 4]]     # b also changed!
```

---

## Copying mutable objects 

Instead, we need the `deepcopy()` method from the `copy` module:

```python
import copy

a = [[1, 2], [3, 4]]
b = copy.deepcopy(a)

a[0][1] = 10

a
#> [[1, 10], [3, 4]]
b
#> [[1, 2], [3, 4]]      # b has not changed
```

---
class: center middle inverse

# Errors and error handling (4.5)

-----

<br> <br> <br> <br>

---

## Errors and exceptions

Python's error messages contain a lot of useful information,
so it really pays off to pay attention to them.

The two main error types that Python distinguishes between are:

- **Syntax errors**: You have not followed the syntax (= grammar) of Python.

- **Exceptions**: The code is gramatically correct but cannot be executed.

---

## Most common error messages

| Error type       | Explanation
|------------------|--------------
| SyntaxError      | Your code contains a syntax (i.e., grammar) error. <br> Common problems are a missing colon, missing quotes, `=` instead of `==`.
| IndentationError | **Incorrect amount of indentation**, e.g. after `if`, loop, or function definition.
| TypeError        | Variable type does not allow the operation: e.g. adding a number and a string.
| NameError        | Variable does not exist (typo?).
| IndexError       | Item in list does not exist.
| KeyError         | Item in dictionary does not exist.
| IOError          | File that does not exist, or you tried to write to a file that is open only for reading (IO = Input/Output).
| AttributeError   | Attribute or method that does not exist, e.g. `.keys()` on a list.

---

## Handling exceptions

Like almost any programming language (but unlike Bash),
Python will stop execution when it encounters an error:
e.g., in case of an error encountered while running your script,
any code after the error will not be run.

For instance, we may get an error when trying to divide by 0:

```python
y = 16
x = 0

y / x
#> ----------------------------------------------------------
#> ZeroDivisionError Traceback (most recent call last)
#> ----> 1 y / x
#> ZeroDivisionError: float division by zero
```

---

## Handling exceptions (cont.)

Sometimes, you want to allow for an error and carry on,
in which case you are said to "catch" the error.

For example:

```python
try:
    print(y / x)
except:
    print("Cannot divide by 0")
print("I'm done")
```

<br>

--

.content-box-info[
Note that this is doing the opposite of what we were doing in Bash scripts,
where we made sure the program stops when it encounters an error,
which it did not do by default.

The distinction is that in *specific, anticipated cases*,
we actually may want to carry on despite something that is technically an error.
]

---
class: center middle inverse

# Questions?

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Bonus material

-----

<br> <br> <br> <br>

---
class: center middle inverse

# Writing style (4.3)

-----

<br> <br> <br> <br>

---
background-color: #f2f5eb

## Some style  guidelines

- For more, see "[PEP8 - The style guide for Python code](https://pep8.org/)".

- Use four spaces per indentation level; not tabs.

- Split long lines (especially the arguments of functions) using parentheses.

- Separate functions using blank lines.

- Use a new line for each operation:

.pull-left[
Yes:
```python
import scipy
import re
```
]

.pull-left[
No:
```python
import scipy,re
```
]

---
background-color: #f2f5eb

## Some style guidelines (cont.)

- Put spaces around operators (unless using the : to “slice” a list); use
parentheses to clarify the precedence of operators:

.pull-left[
Yes:
```python
a = b ** c
z = [a, b, c, d]
n = n * (n - 1)
k = (a * b) + (c * d)
my_list = a_list[1:n]
```
]

.pull-right[
No:
```python
a=b**c
z=[ a,b,c,d ]
n =n*(n-1)
k=a*b+c*d
my_list = a_list[1 : n]
```
]

---
background-color: #f2f5eb

## Some style guidelines (cont.)

- **Use lots of comments and "docstrings"!**

  Python allows you to write short comments (starting with # ) and
  docstrings (multiline comments flanked by triple double quotes, """ ):
  
  - Use docstrings to document *how to use the code*.
    What does this function do? What is the input/output?
  
  - Use short comments for explaining difficult passages in the code,
    that is, to document how the code works.


<br>

- **Naming conventions:**

  - Variables and functions in lowercase, "snake case": `my_var`, `my_fun`.
  
  - **Constants** in uppercase: `SOME_CONSTANT`.
  
  - Use descriptive names: `count_pairs`, `body_mass`, `cell_volume`.
  
  - Verbs for functions and nouns for objects.

